### 2021-01-07 manuscript.tex
first

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Preamble

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Page Settings
\documentclass[10pt]{report}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[letterpaper, margin=1in, left=1.5in]{geometry}

% Bibliography
\usepackage{biblatex}
\addbibresource{ref/thesis.bib}

% Packages and some formatting
\usepackage{titlesec,hyperref,booktabs,relsize,setspace,float,verbatim,caption,subcaption,graphicx}
\hypersetup{colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue}
%\hypersetup{hidelinks}
\titleformat{\chapter}[hang]
{\normalfont\huge\bfseries}{\chaptertitlename\ \thechapter:}{1em}{}
\doublespacing{}
\setcounter{tocdepth}{1}

% Identification
\title{}
\author{Christopher T. Tastad}
\date{day month year}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Introductory Material

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Title Page
\thispagestyle{empty}
\begin{center}
    \vspace*{1cm}

    \Huge
    \textbf{Refactoring the\\Candidate Cancer Gene Database}

    \vspace{1cm}
    \Large
    A Thesis\\
    SUBMITTED TO THE FACULTY OF THE\\
    UNIVERSITY OF MINNESOTA\\
    BY\\

    \vspace{1.5cm}

    \LARGE
    \textbf{Christopher T. Tastad}

    \vspace{3cm}

    \Large
    IN PARTIAL FULFILLMENT OF THE REQUIREMENTS\\
    FOR THE DEGREE OF\\
    MASTER OF SCIENCE\\

    \vspace{1cm}

    Adviser: Timothy K. Starr, PhD\\

    [full month and year of degree conferral]

\end{center}

% Copyright Page
\newpage
\thispagestyle{empty}

MIT License

Copyright (c) 2019 University of Minnesota

Permission is hereby granted, free of charge, to any person obtaining a copy of this software and associated documentation files (the ``Software''), to deal in the Software without restriction, including without limitation the rights to use, copy, modify, merge, publish, distribute, sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED ``AS IS'', WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT\@. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE\@.

%% Acknowledgements
%\newpage
\pagenumbering{roman}
%
%acknowledgements
%
%% Dedication
%\newpage
%
%dedication
%
%% Abstract
%\newpage
%
%abstract

\tableofcontents
\addcontentsline{toc}{chapter}{List of Tables}
\listoftables
\addcontentsline{toc}{chapter}{List of Figures}
\listoffigures

\newpage

\pagenumbering{arabic}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Introduction

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Introduction}
Refactoring is intended to improve the design, structure, and/or implementation of the software (its non-functional attributes), while preserving its functionality. Potential advantages of refactoring may include improved code readability and reduced complexity; these can improve the source code's maintainability and create a simpler, cleaner, or more expressive internal architecture or object model to improve extensibility. Another potential goal for refactoring is improved performance; software engineers face an ongoing challenge to write programs that perform faster or use less memory\cite{Kerievsky2004}

\subsection{the challenge}

field of bio has ever increasing issues in data management and doesn't apply tech solutions very effectively. Ability to move past code to data manipulation and presentation. Deployment and the integration of common programming practices

In many ways, the field of Bioinformatics exists as the result of a problem rather than a discovery, as with other scientific fields. The nature of the challenges have evolved from a growth in the composition of primary data and a lacking compensatory set of technical skills to evaluate it. With the bottle neck that exists at the center of this problem, previously simple notions of collating, analyzing, and visualizing results no longer scale. As a result, a measurable area of focus is not just in the novel discovery of Biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions.

\subsection{the tech has evolved}

Transitioning (implementing) existing solutions with new paradigms.  Improved data management (manipulation, collation, visualization).  Web front-end tech (web dev).  Technologies that enable reformed development (javascript, markdown, rstudio engine). Increased accessibility to non-tech users.

\subsection{applications of the tech in bio, specifically the CCGD}
gene screens in cancer bio~\cite{Clark2019}

Cancer gene screening is a hallmark case for this exploration as it lives at the nexus of being an area of high impact research and a process that can produce a deluge of raw data. In tandem, the state of web front ends has advanced in a manner that more completely allows~\cite{10.1093/nar/gkz1161}
In this work, we present a small but emblematic case that illustrates these shifting paradigms. With these incremental changeovers, no mater the impact, the field takes on a new character in the management of its growing data challenges. The CCGD as a concept set out to tackle some of these original problems of filtering handling large quantities of data. We expand on this by re-implementing the same achievement through a modern implementation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Body

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Body}

\section{Methods}

\subsection{Project Requirements}
This project's goal was ultimately to upgrade the existing CCGD\@. In setting out to do this, we established several requirements that took into consideration the opportunity to seize a routine set of server transitions to implement improved functionality. Specifically, the central output of the service should be retained in the eyes of the end user. Beyond this, attempting to fully reconstruct the backend provided for significant improvements (see Table~\ref{table:projectReqs}). These goals more broadly were: upgrade the existing server build to RHEL 7; rewrite web interface in a more modern framework; streamline the central table construction process using Rshiny; implement improved ability to manipulate table by product owner; employ improved software development best practices; generally seek opportunities for codebase and resource utilization improvements. Ultimately, all of these taken together were intended to serve some form of modernization and simplification.

\begin{table}[H]
    \caption{Project Requirements}\label{table:projectReqs}
    \relsize{-2}
    \addtolength\tabcolsep{-0.2em}
    \begin{tabular}{llllp{7cm}}
        \toprule
        &Feature&Goal&Framework&Description\\
        \midrule
        1&server OS&upgrade server OS&RHEL7&Transition of architecture for public server host. This is required by University OIT due to end of life schedule for RHEL6.\\
        \midrule
        2&web front-end&rewrite web interface&Rmarkdown&Improvements to the web interface written in a modern, simplified language. This will improved access to content creation and allow for automation in site rendering.\\
        \midrule
        3&table build&rewrite table build&Rshiny&Rshiny offers a dramatic improvement to replace the existing process by merging the table build back-end with a modern web display of the app interface.\\
        \midrule
        4&table update&improved admin controls&BASH/R&Existing app version confined some content controls to app author, preventing contributions by app owner.\\
        \midrule
        5&version control&implement best practices&git/docs&No version control was used in the original development of the app. This and other documentation practices were expanded in the rewrite\\
        \midrule
        6&n/a&resource improvements&codebase&Due to the lack of some best practices, there were many opportunities to make impactful resource improvements.\\
        \bottomrule
    \end{tabular}
    \addtolength\tabcolsep{+0.2em}
    \relsize{+2}
\end{table}

\subsection{Server Setup}
At it's core, the OS transition from RHEL 6 to 7 was the original project requirement that spurred this work. Existing hosting for the CCGD was owned by university OIT who established an end of life date for the current server OS\@. As a result, the bulk of core architecture implementation followed a narrow prescription for what was termed a ``Fully Managed'' RHEL7 install as detailed by the OIT-LPT Public-Docs~\cite{OIT}. Contrary to the implications of this installation name, a Fully Managed administration carried a mixture of privileges and management roles that were shared between OIT and our group. This was accomplished by leveraging a chef admin paradigm that was allowed limited sudo privileges within the hosted ecosystem. Importantly, items that extend from the core of server administration such as virtual machine creation, backups, and core security were handled by OIT\@. This left our group with the more focused task of administering exclusively our application.

[to what level of detail should i describe the server setup]

Apache

\subsection{Web Framework}
The upgraded build of the CCGD employed a completely different front-end architecture centralized around R web services environment. Built with dynamic modularity and data transportability in mind, the Rmarkdown framework offers the ability to render several independent markdown files into a unified website. We applied this feature set to produce our general site structure and content as it allowed for a massive simplification in web design. While following the existing site map, we recreated the raw content of the existing application as independent markdown files. Within the Rmarkdown format guide. The keystone feature of the Rmarkdown framework that allows for this unified rendering is the use of a YAML configuration header (see Figure~\ref{fig:siteYaml}). The \texttt{rmarkdown::render\_site} function leverages this configuration file to align and sequentially render content and styling in tandem for a complete site map [citeme rmarkdown docs]. Additionally, individual markdown pages employ this same configuration paradigm for page-specific parameters and formatting guidelines (see Figure~\ref{fig:pageHeader}).

\begin{figure}[H]
    \includegraphics[width=\linewidth]{figs/rmarkdownflow.png}
    \caption{[TEMP] Rmarkdown Rendering Flow}\label{fig:renderFlow}
\end{figure}

[render diagram https://rmarkdown.rstudio.com/lesson-2.html]

responsive web design

An important distinction in summarizing the implementation of web framework is to point out that the Rshiny app, which contains the gene table, is assembled and hosted as a function separate from the general site rendering. More specifically to the web design, the shiny app is hosted at the common public repository \href{https://ctastad.shinyapps.io/table_app/}{\texttt{shinyapps.io}} and rendered in an inline iframe on the site search page. In the eyes of the user this presents a seamless transition, but the functional contrast permits for differential rendering between the two. This allows for the continuous integration of external sources that feed the table while eliminating the need to regularly render the static site content.

\begin{figure}
\centering
    \addtolength\tabcolsep{-0.2em}
    \relsize{-2}
\begin{subfigure}[b]{.5\linewidth}
    \centering
\begin{verbatim}
name: "Candidate Cancer Gene Database"
navbar:
    title: "CCGD"
    left:
        - text: "Home"
          href: index.html
        - text: "Search"
          href: search.html
        - text: "Help"
          href: help.html
        - text: "References"
          href: references.html
        - text: "Contact"
          href: contact.html
\end{verbatim}
\caption{\_site.yml: file for site assembly}\label{fig:siteYaml}
\end{subfigure}%
\begin{subfigure}[b]{.5\textwidth}
\centering
\begin{verbatim}
---
title: "Candidate Cancer Gene Database"
bibliography: refs/ccgd_paper.bib
nocite: '@*'
output:
    html_document:
        includes:
            in_header: "styles/favicon.html"
            after_body: "styles/footer.html"
        css: styles/styles.css
        theme: readable
---
\end{verbatim}
\subcaption{index.Rmd individual page header}\label{fig:pageHeader}
\end{subfigure}
\caption{YAML Configuration Code Segments}\label{fig:yamls}
    \addtolength\tabcolsep{+0.2em}
    \relsize{+2}
\end{figure}

\subsection{Rshiny App}
In order to replace the existing CCGD search functionality our group employed the Rshiny framework. Simply put, the shiny app that now enables this central purpose is an interactive web interface for a single table. Several changes were made to the existing integration pipeline in order to achieve this simplicity.

The structure of a general shiny app follows a highly reduced and consistent paradigm, and ours is no different. The broad format of this work flow is \texttt{ui + server $\rightarrow$ deployment}. The server specifically has a discrete composition made up of reactive constructs of inputs, expressions, and outputs that shape the app's reactive graph, which forms, more accurately, the discrete operation of the application. The assembly and movement of this graph is both emblematic and practical in displaying the modularity and improved resource utilization of the reactive elements. Without belaboring the description of reactive programming, this paradigm employs a declarative programming style that allows for lazy code execution~\cite{Wickham}. What this amounts to in application is for the developer to define loose controls at a higher level while allowing the code to passively fill these requirements. This translates to greater flexibility to both the developer and the machine as resource use can be reduced through modularity in code execution.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{figs/g238.png}
    \caption{table\_app/app.R Reactive Graph}\label{fig:reactiveGraph}
\end{figure}

Within the scope of the \href{https://github.com/ctastad/ccgd/blob/master/table_app/app.R}{\texttt{table\_app/app.R}} server function we had a relatively simple goal of defining variable filters across the CCGD data table. Taking advantage of Rshiny's modularity, we created a two state path for the imported data object that allowed for both a variable-specific filter step along with a static, unchanged display of the table (see Figure~\ref{fig:reactiveGraph}). The advantage in doing this is that the table is displayed immediately on page load without the need to initiate a query. Additionally, the full, unfiltered table is available from the \texttt{downloadData} output at this state as well. For processing going down the filtered path, independent variable inputs are available to the user which are passed through two expression elements. In selecting a variable filter the \texttt{[Vars].values} expression applies a syntactic subsetting of the variable's data list which is then past to the central \texttt{filter.search} where the broader subsetting is carried to the rest of the table. The \texttt{searchTable} function takes the output of the filter and applies several text mutations to add external linking to several associated resources.

A small but fundamental element of our shiny app design was the use of the \texttt{renderDataTable} function in generating the \texttt{searchTable} output. The prior iteration of the CCGD architecture employed a relatively disproportionate MySQL database to store and serve numerous component elements that made up the table. Instead, the new implementation of our app now leverages the server-side processing capability of the DataTables JavaScript library. More specifically, we set parameters to create a reactive version of our function which allows for paging, searching, filtering, and sorting in real-time within the shiny server infrastructure [citeme DataTables]. Through this function alone we established new server paradigm that eliminated the MySQL database.

\subsection{Data Integration Process}

\subsection{Best Practices}
Some features of the software best practices from the previous version of the CCGD were done quite well such as the level of detail in the documentation. At the same time, many other important areas were missing in the development and maintenance of the application. Most prominent among these were the lack of any degree of version control and relatively inconsistent design principles. As a result, our group employed or improved upon several methods in an effort to elevate the quality, usability, and transportability of our codebase and application~\cite{Fincham2011}.

Throughout the course of the project we utilized a private github code repository for hosting and change tracking. This work has since been published on a public repository and assigned a DOI number through Zenodo~\cite{Tastad}. Furthering the spirit of project integrity, we expanded on the nature of the documentation for the user and, more prominently, the application owner. Specifically, documentation that was previously written in a Microsoft Word file was now recreated in markdown, and a notable tone and detail shift was made in serving the information transportability between administrators.

In approaching the process of development, we applied a philosophy of tidy code creation to both syntax and design. A supporting method in creating a more elegant code syntax was the use of R universe tools, tidyverse and Rmarkdown. Both offer a coding grammar that is centered around accessibility and simplification of high-level language organization~\cite{Wickham2019}. Furthering this simplification, we sought opportunities to deduplicate and extract subroutines in the table integration process in an effort to streamline resource constraints. Finally, another notable area for improvement we sought was in disk space utilization. The existing integration process relied on the continuous presence of what were several intermediary files that occupied a sizable disk capacity. To combat this we specifically reworked several data transformation steps to allow for the post-process elimination of these temporary data objects.

\section{Results}

\subsection{Architecture Upgrades}
previous installation used traditional lamp stack

Codebase reduction.

Need for sql database eliminated.

Interface improved. Table content contained within single page.

Following the goals set out in our project requirements, we chose Rmarkdown as a replacement framework to develop the application's web front-end. Previously, the web interface for the application had been developed using Adobe Dreamweaver~\cite{Abbott2015}. For the time Dreamweaver served the purpose of simplifying the process of interface construction and web development, yet we hoped to extend the theme of simplified content creation into a modern framework. Further, utilizing a language such as Rmarkdown almost fully carries the development process into a state of pure content creation.

The original version of the CCGD front-end was developed using Dreamweaver~\cite{Abbott2015}. While this may have suitable at the time, several generations of web development have brought the field to be able to move beyond the process of design and focus more readily on content.

In an effort to reduce code complexity and improve access to content creation, an engineering choice was made to redesign the web front-end in Rmarkdown. While an approach such as this can reduce certain feature availability and overall design flexibility, a conscious priority was placed on the content of the page over the web functionality. In the spirit of our project goals,\ldots A framework such as Rmarkdown provides a development paradigm that dramatically shifts the effort of creation into that content rather than the creation broader page architecture [citeme rmarkdown docs]. The work of assembling that architecture is largely done for the user in leveraging

\subsection{Resource Utilization Improvements}
space constraints eliminated. Server resource reqs reduced overall.

\subsection{Administration Streamlined}
server admin and app controls streamlined. Site render newly automated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Conclusions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Conclusions}

Best practices. Discuss open source.

\addcontentsline{toc}{chapter}{Bibliography}
\printbibliography{}

\end{document}

### 2021-01-12 manuscript.tex
added space tree fig; updated file tree fig
resolved seafile conflict
improved formatting consistency across figs; add color reference to file_tree

@@
 \usepackage{titlesec,hyperref,booktabs,relsize,setspace,float,verbatim,caption,subcaption,graphicx}
-\hypersetup{colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue}
-%\hypersetup{hidelinks}
+%\hypersetup{colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue}
+\hypersetup{hidelinks}
 \titleformat{\chapter}[hang]
@@
 \begin{figure}[H]
-    \includegraphics[width=\linewidth]{figs/rmarkdownflow.png}
-    \caption{[TEMP] Rmarkdown Rendering Flow}\label{fig:renderFlow}
+    \includegraphics[width=\linewidth]{fig/rmarkdown_flow.png}
+    \caption{Rmarkdown Rendering Flow}\label{fig:renderFlow}
 \end{figure}
@@
 \end{subfigure}
-\caption{YAML Configuration Code Segments}\label{fig:yamls}
+\caption{YAML Configuration Segments}\label{fig:yamls}
     \addtolength\tabcolsep{+0.2em}
@@
     \centering
-    \includegraphics[width=\textwidth]{figs/g238.png}
+    \includegraphics[width=\textwidth]{fig/reactive_graph.png}
     \caption{table\_app/app.R Reactive Graph}\label{fig:reactiveGraph}
@@
 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{fig/file_tree.png}
+    \caption{Project directory file tree comparison}\label{fig:fileTree}
+\end{figure}
+
 Following the goals set out in our project requirements, we chose Rmarkdown as a replacement framework to develop the application's web front-end.
@@
 space constraints eliminated. Server resource reqs reduced overall.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{fig/space_tree.png}
+    \caption{Treemap of Disk Use Between App Versions}\label{fig:spaceTree}
+\end{figure}
+
+Ref crossdirstat
 

### 2021-01-13 manuscript.tex
changed rendering flow fig to represent multiple files
create parse_flow fig; made other formatting updates across all figs

@@
 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{fig/parse_flow.png}
+    \caption{Project directory file tree comparison}\label{fig:parseFlow}
+\end{figure}
+
 \subsection{Data Integration Process}
@@
 Throughout the course of the project we utilized a private github code repository for hosting and change tracking.
-This work has since been published on a public repository and assigned a DOI number through Zenodo~\cite{Tastad}.
+This work has since been published on a public repository and assigned a DOI number through \href{https://zenodo.org/record/4422027}{Zenodo}.
 Furthering the spirit of project integrity, we expanded on the nature of the documentation for the user and, more prominently, the application owner.

### 2021-01-19 manuscript.tex
created yaml figs as images; improved fig description copy

@@
 
-% Packages and some formatting
+% Packages
 \usepackage{titlesec,hyperref,booktabs,relsize,setspace,float,verbatim,caption,subcaption,graphicx}
@@
 {\normalfont\huge\bfseries}{\chaptertitlename\ \thechapter:}{1em}{}
+\setcounter{tocdepth}{1}
+
+% Formatting
 \doublespacing{}
-\setcounter{tocdepth}{1}
+%\setlength{\textfloatsep}{30pt plus 2.0pt minus 4.0pt}
 
@@
     \includegraphics[width=\linewidth]{fig/rmarkdown_flow.png}
-    \caption{Rmarkdown Rendering Flow}\label{fig:renderFlow}
+    \caption[Rmarkdown Rendering Flow]{\textbf{Rmarkdown Rendering Flow} The core web framework was created using a stitch process composed of two document conversion steps, which are both contained in the single \texttt{rmarkdow::render\_site} command.
+These two transformations are done by the R-specific knitr package and then the open-source pandoc tool.
+The first phase holds less relevance in particular as its role is to execute R code chunks, of which there are few in our site content.
+The markdown output from this step is seamlessly injected into a pandoc conversion which offers a large variety in document output.
+Specific to our case, the pandoc conversion paired with a companion yaml configuration file allows for multidocument markdown-to-html rendering which is assembled as a complete site stack.
+As a single process, this function generates an integrated, modern web front-end from content written in a light-weight syntax with almost none of the typical complexity of full-scale web page creation.}\label{fig:renderFlow}
 \end{figure}
-
-[render diagram https://rmarkdown.rstudio.com/lesson-2.html]
 
@@
 \centering
-    \addtolength\tabcolsep{-0.2em}
-    \relsize{-2}
-\begin{subfigure}[b]{.5\linewidth}
+\begin{subfigure}[t]{.4\textwidth}
     \centering
-\begin{verbatim}
-name: "Candidate Cancer Gene Database"
-navbar:
-    title: "CCGD"
-    left:
-        - text: "Home"
-          href: index.html
-        - text: "Search"
-          href: search.html
-        - text: "Help"
-          href: help.html
-        - text: "References"
-          href: references.html
-        - text: "Contact"
-          href: contact.html
-\end{verbatim}
+    \includegraphics[width=\textwidth]{fig/page_header.png}
+\subcaption{index.Rmd individual page header}\label{fig:pageHeader}
+\end{subfigure}%
+\hspace{5em}
+\begin{subfigure}[t]{.4\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{fig/site_header.png}
 \caption{\_site.yml: file for site assembly}\label{fig:siteYaml}
-\end{subfigure}%
-\begin{subfigure}[b]{.5\textwidth}
-\centering
-\begin{verbatim}
----
-title: "Candidate Cancer Gene Database"
-bibliography: refs/ccgd_paper.bib
-nocite: '@*'
-output:
-    html_document:
-        includes:
-            in_header: "styles/favicon.html"
-            after_body: "styles/footer.html"
-        css: styles/styles.css
-        theme: readable
----
-\end{verbatim}
-\subcaption{index.Rmd individual page header}\label{fig:pageHeader}
 \end{subfigure}
-\caption{YAML Configuration Segments}\label{fig:yamls}
-    \addtolength\tabcolsep{+0.2em}
-    \relsize{+2}
+\caption[YAML Configuration Segments]{\textbf{YAML Configuration Segments} Pictured above are the two forms of complete yaml configuration headers utilized in Rmarkdown rendering.
+The central takeaway from this illustration is to recognize the heavy lifting performed by these relatively small code segments.
+Figure~\ref{fig:pageHeader} provides an example of the header found in every~.Rmd file that make-up the respective pages of the site.
+These headers provide primarily formatting configuration for the rendering process but include broader functions as well.
+The header can be as simple as listing a title and pandoc output type or include LaTex bibliography functionality and read-in style sheets like the one shown.
+Either way, a few lines of configuration generate wildly different outputs.
+Figure~\ref{fig:siteYaml} presents the yaml file which serves to assemble independent Rmarkdown files into a joined website as described in figure~\ref{fig:renderFlow}.
+There is less configuration here as the only real purpose is to define the site tree for the \texttt{rmarkdown::render\_site} process.}\label{fig:yamls}
 \end{figure}
@@
     \includegraphics[width=\textwidth]{fig/reactive_graph.png}
-    \caption{table\_app/app.R Reactive Graph}\label{fig:reactiveGraph}
+    \caption[\texttt{table\_app/app.R} Reactive Graph]{\textbf{\texttt{table\_app/app.R} Reactive Graph} An Rshiny app design can be defined by its reactive graph.
+The graph above displays the process flow of our app by highlighting modularity in the data streams around the table filtering process.
+The reactive web design is intended to allow for the automatic implementation of changed states applied through the alternate paths of the graph.
+This allows for the passive display of a static representation of the table while allowing for dynamic adjustment to filtering inputs through the \texttt{[Var] $\rightarrow$ [Var].values} path.
+Additionally, this declarative model provides for an interactive table array while only drawing computation when the appropriate event is called.}\label{fig:reactiveGraph}
 \end{figure}
@@
     \includegraphics[width=\textwidth]{fig/parse_flow.png}
-    \caption{Project directory file tree comparison}\label{fig:parseFlow}
+    \caption[Server-side Processing Flow]{\textbf{Server-side Processing Flow} This diagram illustrates a small but fundamental change applied to the application table processing through the lens of the server-client relationship.
+The ``old'' flow shows the prior table processing procedure, which is highly typical of a classic Linux-Apache-MySql-PHP (LAMP) stack website.
+Inputs and calls made by the browser occur from and within a client-side rendering engine which delivers a static request to the server-side SQL database.
+The database returns parsed XML to the browser that is rendered and displayed.
+Our fundamental shift was to move this data processing entirely to JavaScript.
+Typically, this would result in client-side processing that would prohibit large tables like ours.
+Instead, we leveraged a server-side engine to handle asynchronous AJAX requests which are returned with parsed JSON, shown in the ``new'' path.
+This processing flow allows for the use of a single~.csv file as the data source in table presentation, ordering, and pagination, ultimately eliminating the need for a SQL architecture.}\label{fig:parseFlow}
 \end{figure}
@@
     \includegraphics[width=\textwidth]{fig/file_tree.png}
-    \caption{Project directory file tree comparison}\label{fig:fileTree}
+    \caption[Project Directory File Tree Comparison]{\textbf{Project Directory File Tree Comparison} The above figure offers a qualitative assessment of an element of technical debt that was improved in refactoring.
+Shown are the file trees of the site directories of the new and old projects.
+What is illustrated is a dramatic improvement in the tree size and complexity through the elimination of a large degree of content, log, and style files.
+This is accentuated even forther when comparing the source code.
+This was accomplished both in the intentional design of eliminating content redundancy and in the streamlined structure of the Rmarkdown rendering process.
+This allowed for the elimination of several dimensions of broader web design elements.}\label{fig:fileTree}
 \end{figure}
@@
     \includegraphics[width=\textwidth]{fig/space_tree.png}
-    \caption{Treemap of Disk Use Between App Versions}\label{fig:spaceTree}
+    \caption{Disk Use Comparison}\label{fig:spaceTree}
 \end{figure}

### 2021-01-19 manuscript.tex
added fig description for disk use comparison

@@
 
+First, changes to the backend process allowed for the elimination or significant reduction of several large, persistent data files making them transient through asynchronous update.
+(The largest of these are highlighted with asterisk.) Second, the major architecture choice to handoff server-side table processing to a javascript engine instead of MySQL allowed for the elimination of the 10GB database.
+Last,
+
+sql elimination
+
+persistent data files reduced
+
+build table process constructed to make more files dynamic and transient
+
 \begin{figure}[H]
@@
     \includegraphics[width=\textwidth]{fig/space_tree.png}
-    \caption{Disk Use Comparison}\label{fig:spaceTree}
+\caption[Disk Use Comparison]{\textbf{Disk Use Comparison} shows a treemap which illustrates the substantial difference in disk utilization between CCGD versions.
+The original to-scale map on the left gives an exploded view distinguishing the updated app treemap in the bottom right corner with all remaining elements of the map belonging to the old version.
+This visual comparison emphasizes the [10x] space decrease made during the upgrade.
+The fruits of several design choices are present in this space reduction.
+The most notable impacts of these choices are several massive files eliminated or substantially reduced (noted by asterisk).
+Notably, the dump.sql file shown represented the disproportionate size of the database and space payoff earned by eliminating it.
+Also, it can be seen that the persistent cosmic.txt data file is present in both versions.
+Improvements made in post-process trimming of files like this also produced substantial space returns.
+Last, the rebuilt table build process transitioned several data objects shown from persistent to transient.}\label{fig:spaceTree}
 \end{figure}

### 2021-01-20 manuscript.tex
?

@@
 These two transformations are done by the R-specific knitr package and then the open-source pandoc tool.
-The first phase holds less relevance in particular as its role is to execute R code chunks, of which there are few in our site content.
+The first phase holds less relevance in particular as its role is to execute R code chunks, of which there are few in our page content.
 The markdown output from this step is seamlessly injected into a pandoc conversion which offers a large variety in document output.
@@
 This was accomplished both in the intentional design of eliminating content redundancy and in the streamlined structure of the Rmarkdown rendering process.
-This allowed for the elimination of several dimensions of broader web design elements.}\label{fig:fileTree}
+This allowed for the elimination of several dimensions of broader web design and table build data elements.}\label{fig:fileTree}
 \end{figure}
@@
 \subsection{Resource Utilization Improvements}
-space constraints eliminated. Server resource reqs reduced overall.
+Space constraints eliminated. Server resource reqs reduced overall.
+
+Interface now streamlined to single page instead of a bunch of pages.
 

### 2021-02-04 manuscript.tex
adjusted section order for better logical progression; incorporated subroutine figure
noticed error in table script relationship

@@
 
+\subsection{Data Integration Process}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{fig/subroutine_flow.png}
+    \caption[Subroutine flow]{\textbf{Subroutine flow} This diagram offers an overview of the actions and relationships within the executable scripts of the CCGD.
+Prior functionality was more effectively delineated into subroutines for better modularity.
+Also shown is the added functionality brought by the rebuilt data integration process.
+A hallmark of the difference between this new and old process is the use of R and shiny as the central frameworks to build the table as implemented in the build\_table.R and app.R scripts.
+It can also be seen that the integration process is automated with a cron job among other maintenance tasks.
+Last, the ccgd\_upload.sh and knit\_site.R scripts were created to allow for introduction of new table data or the manual execution of any of the app subroutines.}\label{fig:subroutine}
+\end{figure}
+
 \subsection{Rshiny App}
@@
 \end{figure}
-
-\subsection{Data Integration Process}
 

### 2021-02-08 manuscript.tex
write data integration process

@@
 
+Improvements to the data integration process were accomplished using a more simplified framework to eliminate redundancy and provide a more standardized schema.
+The prior implementation employed some tools that are common place, but they were implemented in a fashion that lacked simplicity at times.
+Also, certain proprietary widgets presented functionality and compatibility choke points that generated road blocks to maintenance.
+The most substantial of the changes that addressed these issues was the transition to use of R as the data manipulation toolset.
+More specifically, the~\texttt{build\_table.R} script established a lean ETL process that took existing table data and external reference data to perform a transformation which collated old and new data.
+The product of this process could then be delivered to the R-based shiny app deployment generated by the~\texttt{app.R} script.
+
+Redundancy reduction was acheived within scipt functionality and storage utilization.
+Dedicated script actions that performed house keeping functions, as seen in the~\texttt{build\_table.sh} and~\texttt{backup.sh} (see Figure~\ref{fig:subroutine}), made specific effort to clear all transient data content that otherwise did not require persistent storage.
+This accounted for a large portion of persistent data that was previously left on the server.
+Additionally, script reduction was improved through a reorganzition of the general subroutines within the application.
+A reassignment of script actions were assigned in a more cohesive manner which allowed for both improved simplicity and modularity.
+This is most effectively seen in the manual control~\texttt{ccge\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app.
+
 \begin{figure}[H]
@@
     \includegraphics[width=\textwidth]{fig/subroutine_flow.png}
-    \caption[Subroutine flow]{\textbf{Subroutine flow} This diagram offers an overview of the actions and relationships within the executable scripts of the CCGD.
+    \caption[Subroutine flow]{\textbf{Subroutine flow} This diagram offers an overview of the actions and relationships within the executable scripts of the CCGD\@.
 Prior functionality was more effectively delineated into subroutines for better modularity.
@@
 
+The rebuilt data integration process represents a core accomplishment of the refactoring project.
+A reconstruction of this process and the frameworks utilized to execute it offered notable improvements
+
 \begin{figure}[H]

### 2021-02-08 manuscript.tex
language progress

@@
 A reassignment of script actions were assigned in a more cohesive manner which allowed for both improved simplicity and modularity.
-This is most effectively seen in the manual control~\texttt{ccge\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app.
+This is most effectively seen in the manual control~\texttt{ccge\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app (see Figure~\ref{fig:subroutine}).
 
@@
 
+Improvements to the core achritecture stand as the central accomplishments in undertaking this project.
+
 \begin{figure}[H]

### 2021-02-09 manuscript.tex
rounded out introduction

@@
 \chapter{Introduction}
-Refactoring is intended to improve the design, structure, and/or implementation of the software (its non-functional attributes), while preserving its functionality.
-Potential advantages of refactoring may include improved code readability and reduced complexity; these can improve the source code's maintainability and create a simpler, cleaner, or more expressive internal architecture or object model to improve extensibility.
-Another potential goal for refactoring is improved performance; software engineers face an ongoing challenge to write programs that perform faster or use less memory\cite{Kerievsky2004}
-
-\subsection{the challenge}
-
-field of bio has ever increasing issues in data management and doesn't apply tech solutions very effectively.
-Ability to move past code to data manipulation and presentation.
-Deployment and the integration of common programming practices
-
 In many ways, the field of Bioinformatics exists as the result of a problem rather than a discovery, as with other scientific fields.
@@
 As a result, a measurable area of focus is not just in the novel discovery of Biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions.
+Further, the crossover of Biological and technical expertise has offerred, at times, a clutterred approach to implementing these solutions resulting in less than desirable engineering.
 
-\subsection{the tech has evolved}
+All while this dynamic has developed in the world of Biology, the technology has rapidly improved.
+A pillar of the field's advances is in the area of web technologies, which now offer frameworks and tools that both accomplish more while requiring less of their developers.
+The classical paradigm of the web has evolved from the LAMP stack (Linux, Apache, MySQL, PHP) to improved frameworks such as advanced Javascript, CSS3, and HTML5 that lay the foundation for modern and flexible web applications.
+These improvements have served to enhance the functionality of applications along with creating a reformed development process that is more intuitive, more accessible, and, ultimately, more powerful.
+These advances offer specific utility to the sciences, and following a comparable progress in application bears consideration.
 
-Transitioning (implementing) existing solutions with new paradigms.
-Improved data management (manipulation, collation, visualization).
-Web front-end tech (web dev).
-Technologies that enable reformed development (javascript, markdown, rstudio engine).
-Increased accessibility to non-tech users.
+While the implementation of new paradigms typically applies to future work, it is essential to look back at opportunities to improve past engineering with updated models through the process of code refactoring.
+At a basic level, refactoring is simply restructing or rewriting the content of a codebase while retaining the functionality of that service.
+This can serve many purposes that all aim to accomplish some range of improvements to the backend design, architecture, or performance.
+Within many engineering goals, this process offers access to items that address routine maintainability to wholesale upgrade.
+At times, this may not just be preferrable but essential in the scope of scheduled updates and general obsolescence, making it a critical process in the scope of software development.
 
-\subsection{applications of the tech in bio, specifically the CCGD}
-gene screens in cancer bio~\cite{Clark2019}
-
-Cancer gene screening is a hallmark case for this exploration as it lives at the nexus of being an area of high impact research and a process that can produce a deluge of raw data.
-In tandem, the state of web front ends has advanced in a manner that more completely allows~\cite{10.1093/nar/gkz1161}
 In this work, we present a small but emblematic case that illustrates these shifting paradigms.
 With these incremental changeovers, no mater the impact, the field takes on a new character in the management of its growing data challenges.
-The CCGD as a concept set out to tackle some of these original problems of filtering handling large quantities of data.
-We expand on this by re-implementing the same achievement through a modern implementation.
+The CCGD as a concept set out to tackle some of the original problems of filtering handling large quantities of data.
+We expand on this by re-implementing the same achievement through a modern paradigm.
+
+%Gene screens in cancer bio~\cite{Clark2019}
+
+%Cancer gene screening is a hallmark case for this exploration as it lives at the nexus of being an area of high impact research and a process that can produce a deluge of raw data.
+In tandem, the state of web front ends has advanced in a manner that more completely allows~\cite{10.1093/nar/gkz1161}
 

### 2021-02-10 manuscript.tex
finalized results outline

@@
 \subsection{Architecture Upgrades}
-previous installation used traditional lamp stack
 
-Codebase reduction.
+Accomplished project goals
+
+Design
+
+Table interface now streamlined to single page
+
+New admin tools streamline app control
+
+Site render automated.
+
+Better url
+
+The rebuilt data integration process
+
+Architecture
+
+implemented modern web platform (front-end and table app) from lamp stack
+
+upgraded server from rhel6
 
@@
 
-Interface improved. Table content contained within single page.
+Performance/resources
 
-The rebuilt data integration process represents a core accomplishment of the refactoring project.
-A reconstruction of this process and the frameworks utilized to execute it offered notable improvements
+Space constraints eliminated.
 
-Improvements to the core achritecture stand as the central accomplishments in undertaking this project.
+Server resource reqs reduced overall.
+
+Data files made dynamic and transient
+
+Codebase reduction.
 
@@
 
-\subsection{Resource Utilization Improvements}
+\subsection{Resource and Design Improvements}
+
+Design, architecture, performance
+
 Space constraints eliminated. Server resource reqs reduced overall.

### 2021-02-10 manuscript.tex
broke out figures in to respective files and converted to pdf over png

@@
 % Packages
+\usepackage{svg}
 \usepackage{titlesec,hyperref,booktabs,relsize,setspace,float,verbatim,caption,subcaption,graphicx}
@@
 As a result, a measurable area of focus is not just in the novel discovery of Biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions.
-Further, the crossover of Biological and technical expertise has offerred, at times, a clutterred approach to implementing these solutions resulting in less than desirable engineering.
+Further, the crossover of Biological and technical expertise has offered, at times, a cluttered approach to implementing these solutions resulting in less than desirable engineering.
 
@@
 A pillar of the field's advances is in the area of web technologies, which now offer frameworks and tools that both accomplish more while requiring less of their developers.
-The classical paradigm of the web has evolved from the LAMP stack (Linux, Apache, MySQL, PHP) to improved frameworks such as advanced Javascript, CSS3, and HTML5 that lay the foundation for modern and flexible web applications.
+The classical paradigm of the web has evolved from the LAMP stack (Linux, Apache, MySQL, PHP) to improved frameworks such as advanced JavaScript, CSS3, and HTML5 that lay the foundation for modern and flexible web applications.
 These improvements have served to enhance the functionality of applications along with creating a reformed development process that is more intuitive, more accessible, and, ultimately, more powerful.
@@
 While the implementation of new paradigms typically applies to future work, it is essential to look back at opportunities to improve past engineering with updated models through the process of code refactoring.
-At a basic level, refactoring is simply restructing or rewriting the content of a codebase while retaining the functionality of that service.
+At a basic level, refactoring is simply restructuring or rewriting the content of a codebase while retaining the functionality of that service.
 This can serve many purposes that all aim to accomplish some range of improvements to the backend design, architecture, or performance.
 Within many engineering goals, this process offers access to items that address routine maintainability to wholesale upgrade.
-At times, this may not just be preferrable but essential in the scope of scheduled updates and general obsolescence, making it a critical process in the scope of software development.
+At times, this may not just be preferable but essential in the scope of scheduled updates and general obsolescence, making it a critical process in the scope of software development.
 
@@
 \begin{figure}[H]
-    \includegraphics[width=\linewidth]{fig/rmarkdown_flow.png}
-    \caption[Rmarkdown Rendering Flow]{\textbf{Rmarkdown Rendering Flow} The core web framework was created using a stitch process composed of two document conversion steps, which are both contained in the single \texttt{rmarkdow::render\_site} command.
-These two transformations are done by the R-specific knitr package and then the open-source pandoc tool.
+    \includegraphics[width=\linewidth]{fig/rmarkdown_flow.pdf}
+    \caption[Rmarkdown Rendering Flow]{\textbf{Rmarkdown Rendering Flow} The core web framework was created using a stitch process composed of two document conversion steps, which are both contained in the single \texttt{rmarkdown::render\_site} command.
+These two transformations are done by the R-specific knitr package and then the open-source Pandoc tool.
 The first phase holds less relevance in particular as its role is to execute R code chunks, of which there are few in our page content.
-The markdown output from this step is seamlessly injected into a pandoc conversion which offers a large variety in document output.
-Specific to our case, the pandoc conversion paired with a companion yaml configuration file allows for multidocument markdown-to-html rendering which is assembled as a complete site stack.
+The markdown output from this step is seamlessly injected into a Pandoc conversion which offers a large variety in document output.
+Specific to our case, the Pandoc conversion paired with a companion YAML configuration file allows for multidocument markdown-to-html rendering which is assembled as a complete site stack.
 As a single process, this function generates an integrated, modern web front-end from content written in a light-weight syntax with almost none of the typical complexity of full-scale web page creation.}\label{fig:renderFlow}
@@
     \centering
-    \includegraphics[width=\textwidth]{fig/page_header.png}
+    \includegraphics[width=\textwidth]{fig/page_header.pdf}
 \subcaption{index.Rmd individual page header}\label{fig:pageHeader}
@@
     \centering
-    \includegraphics[width=\textwidth]{fig/site_header.png}
+    \includegraphics[width=\textwidth]{fig/site_header.pdf}
 \caption{\_site.yml: file for site assembly}\label{fig:siteYaml}
 \end{subfigure}
-\caption[YAML Configuration Segments]{\textbf{YAML Configuration Segments} Pictured above are the two forms of complete yaml configuration headers utilized in Rmarkdown rendering.
+\caption[YAML Configuration Segments]{\textbf{YAML Configuration Segments} Pictured above are the two forms of complete YAML configuration headers utilized in Rmarkdown rendering.
 The central takeaway from this illustration is to recognize the heavy lifting performed by these relatively small code segments.
@@
 These headers provide primarily formatting configuration for the rendering process but include broader functions as well.
-The header can be as simple as listing a title and pandoc output type or include LaTex bibliography functionality and read-in style sheets like the one shown.
+The header can be as simple as listing a title and Pandoc output type or include LaTex bibliography functionality and read-in style sheets like the one shown.
 Either way, a few lines of configuration generate wildly different outputs.
-Figure~\ref{fig:siteYaml} presents the yaml file which serves to assemble independent Rmarkdown files into a joined website as described in figure~\ref{fig:renderFlow}.
+Figure~\ref{fig:siteYaml} presents the YAML file which serves to assemble independent Rmarkdown files into a joined website as described in figure~\ref{fig:renderFlow}.
 There is less configuration here as the only real purpose is to define the site tree for the \texttt{rmarkdown::render\_site} process.}\label{fig:yamls}
@@
 
-Redundancy reduction was acheived within scipt functionality and storage utilization.
+Redundancy reduction was achieved within script functionality and storage utilization.
 Dedicated script actions that performed house keeping functions, as seen in the~\texttt{build\_table.sh} and~\texttt{backup.sh} (see Figure~\ref{fig:subroutine}), made specific effort to clear all transient data content that otherwise did not require persistent storage.
 This accounted for a large portion of persistent data that was previously left on the server.
-Additionally, script reduction was improved through a reorganzition of the general subroutines within the application.
+Additionally, script reduction was improved through a reorganization of the general subroutines within the application.
 A reassignment of script actions were assigned in a more cohesive manner which allowed for both improved simplicity and modularity.
-This is most effectively seen in the manual control~\texttt{ccge\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app (see Figure~\ref{fig:subroutine}).
+This is most effectively seen in the manual control~\texttt{ccgd\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app (see Figure~\ref{fig:subroutine}).
 
@@
     \centering
-    \includegraphics[width=\textwidth]{fig/subroutine_flow.png}
+    \includegraphics[width=\textwidth]{fig/subroutine_flow.pdf}
     \caption[Subroutine flow]{\textbf{Subroutine flow} This diagram offers an overview of the actions and relationships within the executable scripts of the CCGD\@.
@@
     \centering
-    \includegraphics[width=\textwidth]{fig/reactive_graph.png}
+    \includegraphics[width=\textwidth]{fig/reactive_graph.pdf}
     \caption[\texttt{table\_app/app.R} Reactive Graph]{\textbf{\texttt{table\_app/app.R} Reactive Graph} An Rshiny app design can be defined by its reactive graph.
@@
     \centering
-    \includegraphics[width=\textwidth]{fig/parse_flow.png}
+    \includegraphics[width=\textwidth]{fig/parse_flow.pdf}
     \caption[Server-side Processing Flow]{\textbf{Server-side Processing Flow} This diagram illustrates a small but fundamental change applied to the application table processing through the lens of the server-client relationship.
-The ``old'' flow shows the prior table processing procedure, which is highly typical of a classic Linux-Apache-MySql-PHP (LAMP) stack website.
+The ``old'' flow shows the prior table processing procedure, which is highly typical of a classic Linux-Apache-MySQL-PHP (LAMP) stack website.
 Inputs and calls made by the browser occur from and within a client-side rendering engine which delivers a static request to the server-side SQL database.
@@
 In approaching the process of development, we applied a philosophy of tidy code creation to both syntax and design.
-A supporting method in creating a more elegant code syntax was the use of R universe tools, tidyverse and Rmarkdown.
+A supporting method in creating a more elegant code syntax was the use of R universe tools, Tidyverse and Rmarkdown.
 Both offer a coding grammar that is centered around accessibility and simplification of high-level language organization~\cite{Wickham2019}.
@@
 
-\subsection{Architecture Upgrades}
-
-Accomplished project goals
+\subsection{Design Improvements}
 
@@
     \centering
-    \includegraphics[width=\textwidth]{fig/file_tree.png}
+    \includegraphics[width=\textwidth]{fig/file_tree.pdf}
     \caption[Project Directory File Tree Comparison]{\textbf{Project Directory File Tree Comparison} The above figure offers a qualitative assessment of an element of technical debt that was improved in refactoring.
@@
 What is illustrated is a dramatic improvement in the tree size and complexity through the elimination of a large degree of content, log, and style files.
-This is accentuated even forther when comparing the source code.
+This is accentuated even further when comparing the source code.
 This was accomplished both in the intentional design of eliminating content redundancy and in the streamlined structure of the Rmarkdown rendering process.
@@
 
-\subsection{Resource and Design Improvements}
-
-Design, architecture, performance
-
-Space constraints eliminated. Server resource reqs reduced overall.
-
-Interface now streamlined to single page instead of a bunch of pages.
-
+\subsection{Architecture Upgrades}
 First, changes to the backend process allowed for the elimination or significant reduction of several large, persistent data files making them transient through asynchronous update.
-(The largest of these are highlighted with asterisk.) Second, the major architecture choice to handoff server-side table processing to a javascript engine instead of MySQL allowed for the elimination of the 10GB database.
+(The largest of these are highlighted with asterisk.) Second, the major architecture choice to hand off server-side table processing to a JavaScript engine instead of MySQL allowed for the elimination of the 10GB database.
 Last,
-
-sql elimination
-
-persistent data files reduced
-
-build table process constructed to make more files dynamic and transient
 
@@
     \centering
-    \includegraphics[width=\textwidth]{fig/space_tree.png}
+    \includegraphics[width=\textwidth]{fig/space_tree.pdf}
 \caption[Disk Use Comparison]{\textbf{Disk Use Comparison} shows a treemap which illustrates the substantial difference in disk utilization between CCGD versions.
@@
 
-\subsection{Administration Streamlined}
-server admin and app controls streamlined. Site render newly automated.
+\subsection{Resources Improvements}
 

### 2021-02-11 manuscript.tex
wrote architecture and design results sections

@@
 
-While the implementation of new paradigms typically applies to future work, it is essential to look back at opportunities to improve past engineering with updated models through the process of code refactoring.
+While the implementation of new models typically applies to future work, it is essential to look back at opportunities to improve past engineering with updated models through the process of code refactoring.
 At a basic level, refactoring is simply restructuring or rewriting the content of a codebase while retaining the functionality of that service.
@@
 The CCGD as a concept set out to tackle some of the original problems of filtering handling large quantities of data.
-We expand on this by re-implementing the same achievement through a modern paradigm.
+We expand on this by re-implementing the same achievement through a modern application.
 
@@
 This left our group with the more focused task of administering exclusively our application.
-
-[to what level of detail should i describe the server setup]
-
-Apache
 
@@
 The assembly and movement of this graph is both emblematic and practical in displaying the modularity and improved resource utilization of the reactive elements.
-Without belaboring the description of reactive programming, this paradigm employs a declarative programming style that allows for lazy code execution~\cite{Wickham}.
+Without belaboring the description of reactive programming, this model employs a declarative programming style that allows for lazy code execution~\cite{Wickham}.
 What this amounts to in application is for the developer to define loose controls at a higher level while allowing the code to passively fill these requirements.
@@
 
-\subsection{Design Improvements}
+\subsection{Architecture Upgrades}
+Following the goals set out in our project requirements, we chose Rmarkdown as a replacement framework to develop the application's web front-end.
+Previously, the web interface for the application had been developed using Adobe Dreamweaver~\cite{Abbott2015} resting on a traditional LAMP stack.
+For the time, a LAMP stack and Dreamweaver served the purpose of simplifying the process of interface construction on a stable and consistent web platform, yet we hoped to extend the theme of simplified content creation into a modern framework.
+Further, utilizing a language such as Rmarkdown almost fully carries the development process into a state of pure content creation.
+Through the Rmarkdown toolset, the work of web design is now largely rendered automatically for the developer.
+Employing the R web scheme established a cornerstone for the project to leverage both enhanced web development as well as making available other key tools in architecture design choices.
 
-Design
+Continuing with those architecture choices, Rshiny stood as a natural segue from the Rmarkdown front-end to assemble the table application itself.
+By upgrading the web paradigm from that LAMP stack used in the prior version, we were able to shed some of the burden that come with that classic architecture in creating the new table application.
+This was further realized by leveraging the Rshiny toolkit.
+A central premise of the Rshiny design philosophy promotes interactive or reactive web applications.
+With that philosophy comes a set of nimble features that are flexible, lightweight, and, still, very powerful.
+Taken together, refactoring the table in Rshiny code allowed us to fully replace the functionality of a SQL-based table with a JavaScript-powered backend that conducts previously resources intensive tasks in real-time.
+This is all furthered by the fact that the dramatically improved ease of development found in Rmarkdown, largely carries into Rshiny as well.
 
-Table interface now streamlined to single page
+\subsection{Design Changes}
+The foundation laid by these architecture changes provided many substantive avenues for improved design.
+Following best practices as well we also developed improved workflows which extended those core structural improvements.
+The most substantive of these gains was found in the complete rebuild performed on the data integration process.
+Given that we established a new framework built on R, it was necessary to extend modifications to the length of the full pipeline.
+In doing so, we also explored improvements in the base process peripheral to the new R code as there were opportunities to improve these workflows.
+As a result, we broke several existing functions into dedicated subroutines for improved modularity in both script execution and data management.
+Additionally, the considerations given to redesigning these script functions led to the development of improved app inputs in the form of more accessible manual controls.
+The convergence of the advances made in these design choices is maybe best illustrated by the site render function.
+A functionality newly created, the \texttt{knit\_site.R} script was only made possible through the features and ease of the Rmarkdown architecture.
+The effort to package this process as a subroutine elevated these mechanisms to the level of being a component of the application rather than just actions performed to conduct development.
 
-New admin tools streamline app control
+The improved web front-end not only offered a needed facelift to the aesthetic appearance of the site, but the updated framework brought many of the elements of the page construction current with existing browser standards.
 
-Site render automated.
+- Table interface now streamlined to single page
+- Better url
 
-Better url
-
-The rebuilt data integration process
-
-Architecture
-
-implemented modern web platform (front-end and table app) from lamp stack
-
-upgraded server from rhel6
-
-Need for sql database eliminated.
+- New admin tools streamline app control
+- Site render automated.
+- The rebuilt data integration process
 
@@
 \end{figure}
-
-Following the goals set out in our project requirements, we chose Rmarkdown as a replacement framework to develop the application's web front-end.
-Previously, the web interface for the application had been developed using Adobe Dreamweaver~\cite{Abbott2015}.
-For the time Dreamweaver served the purpose of simplifying the process of interface construction and web development, yet we hoped to extend the theme of simplified content creation into a modern framework.
-Further, utilizing a language such as Rmarkdown almost fully carries the development process into a state of pure content creation.
-
-The original version of the CCGD front-end was developed using Dreamweaver~\cite{Abbott2015}.
-While this may have suitable at the time, several generations of web development have brought the field to be able to move beyond the process of design and focus more readily on content.
-
-In an effort to reduce code complexity and improve access to content creation, an engineering choice was made to redesign the web front-end in Rmarkdown.
-While an approach such as this can reduce certain feature availability and overall design flexibility, a conscious priority was placed on the content of the page over the web functionality.
-In the spirit of our project goals,\ldots A framework such as Rmarkdown provides a development paradigm that dramatically shifts the effort of creation into that content rather than the creation broader page architecture [citeme rmarkdown docs].
-The work of assembling that architecture is largely done for the user in leveraging
-
-\subsection{Architecture Upgrades}
-First, changes to the backend process allowed for the elimination or significant reduction of several large, persistent data files making them transient through asynchronous update.
-(The largest of these are highlighted with asterisk.) Second, the major architecture choice to hand off server-side table processing to a JavaScript engine instead of MySQL allowed for the elimination of the 10GB database.
-Last,
 

### 2021-02-11 manuscript.tex
wrote resources improvement section; added more fig refs

@@
 \subsection{Data Integration Process}
-
 Improvements to the data integration process were accomplished using a more simplified framework to eliminate redundancy and provide a more standardized schema.
@@
 Further, utilizing a language such as Rmarkdown almost fully carries the development process into a state of pure content creation.
-Through the Rmarkdown toolset, the work of web design is now largely rendered automatically for the developer.
+Through the Rmarkdown toolset, the work of web design is now largely rendered automatically for the developer (see Figure~\ref{fig:renderFlow}).
 Employing the R web scheme established a cornerstone for the project to leverage both enhanced web development as well as making available other key tools in architecture design choices.
@@
 This was further realized by leveraging the Rshiny toolkit.
-A central premise of the Rshiny design philosophy promotes interactive or reactive web applications.
+A central premise of the Rshiny design philosophy promotes interactive or reactive web applications (see Figure~\ref{fig:reactiveGraph}).
 With that philosophy comes a set of nimble features that are flexible, lightweight, and, still, very powerful.
-Taken together, refactoring the table in Rshiny code allowed us to fully replace the functionality of a SQL-based table with a JavaScript-powered backend that conducts previously resources intensive tasks in real-time.
+Taken together, refactoring the table in Rshiny code allowed us to fully replace the functionality of a SQL-based table with a JavaScript-powered backend that conducts previously resources intensive tasks in real-time (see Figure~\ref{fig:parseFlow}).
 This is all furthered by the fact that the dramatically improved ease of development found in Rmarkdown, largely carries into Rshiny as well.
@@
 A functionality newly created, the \texttt{knit\_site.R} script was only made possible through the features and ease of the Rmarkdown architecture.
-The effort to package this process as a subroutine elevated these mechanisms to the level of being a component of the application rather than just actions performed to conduct development.
+The effort to package this process as a subroutine elevated these mechanisms to the level of being a component of the application rather than just actions performed to conduct development (see Figure~\ref{fig:subroutine}).
 
-The improved web front-end not only offered a needed facelift to the aesthetic appearance of the site, but the updated framework brought many of the elements of the page construction current with existing browser standards.
+Along with software design changes, there were several improvements made to the interface and site appearance.
+The upgraded web front-end not only offered a needed facelift to the aesthetic appearance of the site, but the new framework brought many of the elements of the page construction current with existing browser standards.
+While not an original goal established in our project requirements, the many considerations that go into web element compatibility are well served in this update, ultimately reducing risk of unintended obsolescence.
+Minor usability factors were considered as well such as the creation of an improved URL\@.
+The CCGD can now be accessed via~\href{https://z.umn.edu/ccgd}{(z.umn.edu/ccgd)} rather than the much more unruly~\href{http://hst-ccgd-prd-web.oit.umn.edu/}{(http://hst-ccgd-prd-web.oit.umn.edu/)}.
+A more major usability achievement was derived from the structure of the new Rshiny table interface.
+With the combination of the reorganized data integration process, the elimination of the SQL backend, and the front-end chosen in the DataTables library, we were able to assemble the application as a single page.
+This dramatically reduced broader site map complexity and simplified the process of interacting with the same data set for the user overall.
 
-- Table interface now streamlined to single page
-- Better url
+\subsection{Resource Use Improvements}
+Special consideration was given to resource use throughout the refactoring process.
+Generally as a measure of best practice, resource improvements were sought at all times within engineering deliberation.
+Still, a more concrete issue was present with the recent development of space constraints in the existing production version shortly prior to the project start.
+We sought ways to most effectively address this acute issue in the existing design and identified several persistent data files as notable offenders of disk use.
+Notably, several large reference files in the data integration process were being written to disk.
+We assessed these files and identified several candidates for house keeping integration post-process and found ways to compress those that need remain persistent.
+This alone generated approximately a 50\% reduction in existing space use.
+Further dramatic space reduction was realized with the elimination of the SQL database.
+All together, these combined changes took a server instance that sat at approximately 23GB and brought it to 150MB\@ (see Figure~\ref{fig:spaceTree}).
 
-- New admin tools streamline app control
-- Site render automated.
-- The rebuilt data integration process
-
-Performance/resources
-
-Space constraints eliminated.
-
-Server resource reqs reduced overall.
-
-Data files made dynamic and transient
-
-Codebase reduction.
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{fig/file_tree.pdf}
-    \caption[Project Directory File Tree Comparison]{\textbf{Project Directory File Tree Comparison} The above figure offers a qualitative assessment of an element of technical debt that was improved in refactoring.
-Shown are the file trees of the site directories of the new and old projects.
-What is illustrated is a dramatic improvement in the tree size and complexity through the elimination of a large degree of content, log, and style files.
-This is accentuated even further when comparing the source code.
-This was accomplished both in the intentional design of eliminating content redundancy and in the streamlined structure of the Rmarkdown rendering process.
-This allowed for the elimination of several dimensions of broader web design and table build data elements.}\label{fig:fileTree}
-\end{figure}
+Another broad concept we addressed in elevating general efficiency was exploring general concepts of technical debt.
+This idea is addressed in our approach to dealing with space constraints on some level, but it is a problem that can generally permeate an entire codebase.
+An effective illustration of this is given by the qualitative view of the total file tree present in the site directory (see Figure~\ref{fig:fileTree}).
+It can be difficult to pin point the driving factor in what leads to the accumulation of software entropy, but it is easy to see it exists nonetheless.
+By continuing our emphasis on best practices, we were able to leverage the simplified architecture upgrades alongside elimination of some functional redundancy to generate a working code base that spans across a much more tidy tree.
 
@@
 
-\subsection{Resources Improvements}
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{fig/file_tree.pdf}
+    \caption[Project Directory File Tree Comparison]{\textbf{Project Directory File Tree Comparison} The above figure offers a qualitative assessment of an element of technical debt that was improved in refactoring.
+Shown are the file trees of the site directories of the new and old projects.
+What is illustrated is a dramatic improvement in the tree size and complexity through the elimination of a large degree of content, log, and style files.
+This is accentuated even further when comparing the source code.
+This was accomplished both in the intentional design of eliminating content redundancy and in the streamlined structure of the Rmarkdown rendering process.
+This allowed for the elimination of several dimensions of broader web design and table build data elements.}\label{fig:fileTree}
+\end{figure}
 

### 2021-02-12 manuscript.tex
converted bibtex style to zotero and updated cite keys

@@
 Existing hosting for the CCGD was owned by university OIT who established an end of life date for the current server OS\@.
-As a result, the bulk of core architecture implementation followed a narrow prescription for what was termed a ``Fully Managed'' RHEL7 install as detailed by the OIT-LPT Public-Docs~\cite{OIT}.
+As a result, the bulk of core architecture implementation followed a narrow prescription for what was termed a ``Fully Managed'' RHEL7 install as detailed by the OIT-LPT Public-Docs~\cite{oitOITLPTPublicDocsThis}.
 Contrary to the implications of this installation name, a Fully Managed administration carried a mixture of privileges and management roles that were shared between OIT and our group.
@@
 The assembly and movement of this graph is both emblematic and practical in displaying the modularity and improved resource utilization of the reactive elements.
-Without belaboring the description of reactive programming, this model employs a declarative programming style that allows for lazy code execution~\cite{Wickham}.
+Without belaboring the description of reactive programming, this model employs a declarative programming style that allows for lazy code execution~\cite{wickhamMasteringShinyBook}.
 What this amounts to in application is for the developer to define loose controls at a higher level while allowing the code to passively fill these requirements.
@@
 Most prominent among these were the lack of any degree of version control and relatively inconsistent design principles.
-As a result, our group employed or improved upon several methods in an effort to elevate the quality, usability, and transportability of our codebase and application~\cite{Fincham2011}.
+As a result, our group employed or improved upon several methods in an effort to elevate the quality, usability, and transportability of our codebase and application~\cite{finchamSoftwareDevelopmentPractices2011}.
 
@@
 A supporting method in creating a more elegant code syntax was the use of R universe tools, Tidyverse and Rmarkdown.
-Both offer a coding grammar that is centered around accessibility and simplification of high-level language organization~\cite{Wickham2019}.
+Both offer a coding grammar that is centered around accessibility and simplification of high-level language organization~\cite{wickhamWelcomeTidyverse2019}.
 Furthering this simplification, we sought opportunities to deduplicate and extract subroutines in the table integration process in an effort to streamline resource constraints.
@@
 Following the goals set out in our project requirements, we chose Rmarkdown as a replacement framework to develop the application's web front-end.
-Previously, the web interface for the application had been developed using Adobe Dreamweaver~\cite{Abbott2015} resting on a traditional LAMP stack.
+Previously, the web interface for the application had been developed using Adobe Dreamweaver~\cite{abbottCandidateCancerGene2015a} resting on a traditional LAMP stack.
 For the time, a LAMP stack and Dreamweaver served the purpose of simplifying the process of interface construction on a stable and consistent web platform, yet we hoped to extend the theme of simplified content creation into a modern framework.

### 2021-05-07 manuscript.tex
updated table 2.1 feature
idk
created interface fig; updated style for subroutine fig

@@
 The upgraded build of the CCGD employed a completely different front-end architecture centralized around R web services environment.
-Built with dynamic modularity and data transportability in mind, the Rmarkdown framework offers the ability to render several independent markdown files into a unified website.
+Built with dynamic modularity and data portability in mind, the Rmarkdown framework offers the ability to render several independent markdown files into a unified website.
 We applied this feature set to produce our general site structure and content as it allowed for a massive simplification in web design.
@@
 \centering
-\begin{subfigure}[t]{.4\textwidth}
+\begin{subfigure}[t]{.3\textwidth}
     \centering
@@
 \end{subfigure}%
-\hspace{5em}
-\begin{subfigure}[t]{.4\textwidth}
+\hspace{2em}
+\begin{subfigure}[t]{.3\textwidth}
     \centering
@@
 \caption{\_site.yml: file for site assembly}\label{fig:siteYaml}
+\end{subfigure}
+\hspace{2em}
+\begin{subfigure}[t]{.2\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{fig/yaml_illustration.pdf}
+\caption{Illustration of yaml implementation}\label{fig:yamlIllust}
 \end{subfigure}
@@
 The prior implementation employed some tools that are common place, but they were implemented in a fashion that lacked simplicity at times.
-Also, certain proprietary widgets presented functionality and compatibility choke points that generated road blocks to maintenance.
+Also, certain proprietary widgets, such as the need for conversion in and out of the Microsoft Excel format, presented functionality and compatibility choke points that generated road blocks to maintenance.
 The most substantial of the changes that addressed these issues was the transition to use of R as the data manipulation toolset.
-More specifically, the~\texttt{build\_table.R} script established a lean ETL process that took existing table data and external reference data to perform a transformation which collated old and new data.
+More specifically, the~\texttt{build\_table.R} script established a lean extract, transform, load (ETL) process that took the existing table data and external reference data to perform a transformation which collated old and new data.
 The product of this process could then be delivered to the R-based shiny app deployment generated by the~\texttt{app.R} script.
@@
 Additionally, script reduction was improved through a reorganization of the general subroutines within the application.
-A reassignment of script actions were assigned in a more cohesive manner which allowed for both improved simplicity and modularity.
-This is most effectively seen in the manual control~\texttt{ccgd\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app (see Figure~\ref{fig:subroutine}).
+A reassignment of script actions were also implemented in a more cohesive manner which allowed for both improved simplicity and modularity.
+This is most effectively seen in the manual control~\texttt{ccgd\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app (see Figure~\ref{fig:subroutine}) if run manually.
+All of this core functionality was then was automated with scheduled using system cron jobs to sustain routine maintenance in upkeep of the application's table.
 
@@
 Also shown is the added functionality brought by the rebuilt data integration process.
-A hallmark of the difference between this new and old process is the use of R and shiny as the central frameworks to build the table as implemented in the build\_table.R and app.R scripts.
-It can also be seen that the integration process is automated with a cron job among other maintenance tasks.
-Last, the ccgd\_upload.sh and knit\_site.R scripts were created to allow for introduction of new table data or the manual execution of any of the app subroutines.}\label{fig:subroutine}
+A hallmark of the difference between this new and old process is the use of R and shiny as the central frameworks to build the table as implemented in the \texttt{build\_table.R} and \texttt{app.R} scripts.
+It can also be seen that the integration process is automated by the system with a cron job among other maintenance tasks.
+Last, the \texttt{ccgd\_upload.sh} and \texttt{knit\_site.R} scripts were created to allow for introduction of new table data or the manual execution of any of the app subroutines.}\label{fig:subroutine}
 \end{figure}
@@
 A small but fundamental element of our shiny app design was the use of the \texttt{renderDataTable} function in generating the \texttt{searchTable} output.
-The prior iteration of the CCGD architecture employed a relatively disproportionate MySQL database to store and serve numerous component elements that made up the table.
+The prior iteration of the CCGD architecture employed a relatively disproportionate Structured Query Language (SQL) database to store and serve numerous component elements that made up the table.
 Instead, the new implementation of our app now leverages the server-side processing capability of the DataTables JavaScript library.
 More specifically, we set parameters to create a reactive version of our function which allows for paging, searching, filtering, and sorting in real-time within the shiny server infrastructure [citeme DataTables].
-Through this function alone we established new server paradigm that eliminated the MySQL database.
+Through this function alone we established new server paradigm that eliminated the SQL database.
 
@@
 Inputs and calls made by the browser occur from and within a client-side rendering engine which delivers a static request to the server-side SQL database.
-The database returns parsed XML to the browser that is rendered and displayed.
+The database returns content parsed in Extensible Markup Language (XML) to the browser that is rendered and displayed.
 Our fundamental shift was to move this data processing entirely to JavaScript.
 Typically, this would result in client-side processing that would prohibit large tables like ours.
-Instead, we leveraged a server-side engine to handle asynchronous AJAX requests which are returned with parsed JSON, shown in the ``new'' path.
+Instead, we leveraged a server-side engine to handle Asynchronous JavaScript and XML (AJAX) requests which are returned with parsed JavaScript Object Notation (JSON), shown in the ``new'' flow.
 This processing flow allows for the use of a single~.csv file as the data source in table presentation, ordering, and pagination, ultimately eliminating the need for a SQL architecture.}\label{fig:parseFlow}
@@
 Most prominent among these were the lack of any degree of version control and relatively inconsistent design principles.
-As a result, our group employed or improved upon several methods in an effort to elevate the quality, usability, and transportability of our codebase and application~\cite{finchamSoftwareDevelopmentPractices2011}.
+As a result, our group employed or improved upon several methods in an effort to elevate the quality, usability, and portability of our codebase and application~\cite{finchamSoftwareDevelopmentPractices2011}.
 
@@
 Furthering the spirit of project integrity, we expanded on the nature of the documentation for the user and, more prominently, the application owner.
-Specifically, documentation that was previously written in a Microsoft Word file was now recreated in markdown, and a notable tone and detail shift was made in serving the information transportability between administrators.
+Specifically, documentation that was previously written in a Microsoft Word file was now recreated in markdown, and a notable tone and detail shift was made in serving the information portability between administrators.
 
@@
 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{fig/interface.pdf}
+    \caption[Search Interface Improvements]{\textbf{Search Interface Improvements}}\label{fig:interface}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+Mention cosmic connection.
+
 \addcontentsline{toc}{chapter}{Bibliography}

### 2021-05-07 manuscript.tex
added abstract

@@
 
-%% Acknowledgements
-%\newpage
+% Acknowledgements
+\newpage
 \pagenumbering{roman}
-%
-%acknowledgements
-%
-%% Dedication
-%\newpage
-%
-%dedication
-%
-%% Abstract
-%\newpage
-%
-%abstract
+
+acknowledgements
+
+% Dedication
+\newpage
+
+dedication
+
+% Abstract
+\newpage
+
+\section{Abstract}
+Maintenance and upgrades are a fundamental requirement of any software-based tool, and the resources within the field of Bioinformatics are no exception.
+Too often, the combination of self-taught programmers in former Biologists and the broader lack of a software development ecosystem can propagate elevated degrees of software entropy, also known as technical debt.
+It is then a fair assumption to say that this level of entropy is higher than in fields which more consistently employ the skills of professional developers.
+Here, we present the case of the Candidate Cancer Gene Database as a minor instance of taking advantage of an upgrade cycle in an attempt to service some of this technical debt.
+The goals of this project ultimately strived to acheive the implementation of a new, core framework at the base of the data integration backend as well as a more modern and simplified web front-end.
+In all, this represented a complete rebuild of the application while preserving the existing conceptual implementation.
+Our work served to establish a new application with greater structural simplicity, improved resource use, and, most relevant, a code base which will be easier to maintain and upgrade in the future.
 
@@
 Beyond this, attempting to fully reconstruct the backend provided for significant improvements (see Table~\ref{table:projectReqs}).
-These goals more broadly were: upgrade the existing server build to RHEL 7; rewrite web interface in a more modern framework; streamline the central table construction process using Rshiny; implement improved ability to manipulate table by product owner; employ improved software development best practices; generally seek opportunities for codebase and resource utilization improvements.
+These goals more broadly were: upgrade the existing server build to RHEL 7; rewrite web interface in a more modern framework; streamline the central table construction process using Rshiny; implement improved ability to manipulate table and web content by product owner (Tim Starr); employ improved software development best practices; generally seek opportunities for codebase and resource utilization improvements.
 Ultimately, all of these taken together were intended to serve some form of modernization and simplification.
@@
         2&web front-end&rewrite web interface&Rmarkdown&Improvements to the web interface written in a modern, simplified language.
-This will improved access to content creation and allow for automation in site rendering.\\
+This improved access to content creation and allowed for automation in front-end rendering.\\
         \midrule
-        3&table build&rewrite table build&Rshiny&Rshiny offers a dramatic improvement to replace the existing process by merging the table build back-end with a modern web display of the app interface.\\
+        3&table build&rewrite table build&Rshiny&Rshiny offered a dramatic improvement to replace the existing process by merging the table build back-end with a modern web display of the app interface.\\
         \midrule
-        4&table update&improved admin controls&BASH/R&Existing app version confined some content controls to app author, preventing contributions by app owner.\\
+        4&content update&improved admin controls&BASH/R&Old app version confined some content controls to app author, limiting ability to make contributions by product owner (Tim Starr).\\
         \midrule
-        5&version control&implement best practices&git/docs&No version control was used in the original development of the app. This and other documentation practices were expanded in the rewrite\\
+        5&version control&implement best practices&git/docs&No version control was used in the original development of the app. This and other documentation practices were expanded in the rewrite.\\
         \midrule
-        6&n/a&resource improvements&codebase&Due to the lack of some best practices, there were many opportunities to make impactful resource improvements.\\
+        6&multiple&resource improvements&codebase&Due to the lack of some best practices, there were many opportunities to make impactful resource improvements.\\
         \bottomrule
@@
 Contrary to the implications of this installation name, a Fully Managed administration carried a mixture of privileges and management roles that were shared between OIT and our group.
-This was accomplished by leveraging a chef admin paradigm that was allowed limited sudo privileges within the hosted ecosystem.
+This was accomplished by leveraging a chef admin paradigm that allowed limited sudo privileges within the hosted ecosystem.
 Importantly, items that extend from the core of server administration such as virtual machine creation, backups, and core security were handled by OIT\@.
@@
 \begin{figure}[H]
-    \includegraphics[width=\linewidth]{fig/rmarkdown_flow.pdf}
+    \center{}
+    \includegraphics[width=0.8\textwidth]{fig/rmarkdown_flow.pdf}
     \caption[Rmarkdown Rendering Flow]{\textbf{Rmarkdown Rendering Flow} The core web framework was created using a stitch process composed of two document conversion steps, which are both contained in the single \texttt{rmarkdown::render\_site} command.
@@
 \begin{figure}
-\centering
+    \center{}
     \begin{subfigure}[t]{.3\textwidth}
-    \centering
+        \center{}
         \includegraphics[width=\textwidth]{fig/page_header.pdf}
@@
     \begin{subfigure}[t]{.3\textwidth}
-    \centering
+        \center{}
         \includegraphics[width=\textwidth]{fig/site_header.pdf}
@@
     \begin{subfigure}[t]{.2\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{fig/yaml_illustration.pdf}
+        \center{}
+        \includegraphics[width=0.8\textwidth]{fig/yaml_illustration.pdf}
         \caption{Illustration of yaml implementation}\label{fig:yamlIllust}
@@
 \begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{fig/subroutine_flow.pdf}
+    \center{}
+    \includegraphics[width=0.8\textwidth]{fig/subroutine_flow.pdf}
     \caption[Subroutine flow]{\textbf{Subroutine flow} This diagram offers an overview of the actions and relationships within the executable scripts of the CCGD\@.
@@
 \begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{fig/reactive_graph.pdf}
+    \center{}
+    \includegraphics[width=0.8\textwidth]{fig/reactive_graph.pdf}
     \caption[\texttt{table\_app/app.R} Reactive Graph]{\textbf{\texttt{table\_app/app.R} Reactive Graph} An Rshiny app design can be defined by its reactive graph.
@@
 \begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{fig/parse_flow.pdf}
+    \center{}
+    \includegraphics[width=0.8\textwidth]{fig/parse_flow.pdf}
     \caption[Server-side Processing Flow]{\textbf{Server-side Processing Flow} This diagram illustrates a small but fundamental change applied to the application table processing through the lens of the server-client relationship.
@@
 This work has since been published on a public repository and assigned a DOI number through \href{https://zenodo.org/record/4422027}{Zenodo}.
-Furthering the spirit of project integrity, we expanded on the nature of the documentation for the user and, more prominently, the application owner.
+Furthering the spirit of project integrity, we expanded on the nature of the documentation for the user and, more prominently, the product owner.
 Specifically, documentation that was previously written in a Microsoft Word file was now recreated in markdown, and a notable tone and detail shift was made in serving the information portability between administrators.
@@
 \begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{fig/space_tree.pdf}
+    \center{}
+    \includegraphics[width=0.8\textwidth]{fig/space_tree.pdf}
     \caption[Disk Use Comparison]{\textbf{Disk Use Comparison} shows a treemap which illustrates the substantial difference in disk utilization between CCGD versions.
@@
 \begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{fig/file_tree.pdf}
+    \center{}
+    \includegraphics[width=0.8\textwidth]{fig/file_tree.pdf}
     \caption[Project Directory File Tree Comparison]{\textbf{Project Directory File Tree Comparison} The above figure offers a qualitative assessment of an element of technical debt that was improved in refactoring.
@@
 \begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{fig/interface.pdf}
-    \caption[Search Interface Improvements]{\textbf{Search Interface Improvements}}\label{fig:interface}
+    \center{}
+    \includegraphics[width=0.8\textwidth]{fig/interface.pdf}
+    \caption[Search Interface Changes]{\textbf{Search Interface Changes}}\label{fig:interface}
 \end{figure}

### 2021-05-07 manuscript.tex
added conclusion

@@
 Maintenance and upgrades are a fundamental requirement of any software-based tool, and the resources within the field of Bioinformatics are no exception.
-Too often, the combination of self-taught programmers in former Biologists and the broader lack of a software development ecosystem can propagate elevated degrees of software entropy, also known as technical debt.
+Too often, the combination of using self-taught programmers who are typically trained Biologists along with the broader lack of a software development ecosystem can propagate elevated degrees of software entropy, also known as technical debt.
 It is then a fair assumption to say that this level of entropy is higher than in fields which more consistently employ the skills of professional developers.
-Here, we present the case of the Candidate Cancer Gene Database as a minor instance of taking advantage of an upgrade cycle in an attempt to service some of this technical debt.
+Here, we present the case of the Candidate Cancer Gene Database (CCGD, https://z.umn.edu/ccgd) as a minor instance of taking advantage of an upgrade cycle in an attempt to service some of this technical debt.
 The goals of this project ultimately strived to acheive the implementation of a new, core framework at the base of the data integration backend as well as a more modern and simplified web front-end.
@@
     \caption[Project Directory File Tree Comparison]{\textbf{Project Directory File Tree Comparison} The above figure offers a qualitative assessment of an element of technical debt that was improved in refactoring.
-Shown are the file trees of the site directories of the new and old projects.
-What is illustrated is a dramatic improvement in the tree size and complexity through the elimination of a large degree of content, log, and style files.
+Shown are the horizontally oriented file trees of the site directories of the new and old projects.
+What is illustrated is a dramatic improvement in the tree size and complexity through the elimination of a large degree of redudant or unecessary content, log, and style files.
 This is accentuated even further when comparing the source code.
@@
 \chapter{Conclusions}
+This work has given new life to the CCGD and the valuable functionality that it offers in the form of accessible forward genetic cancer screen data.
+Further, it has been a case study in the benefits of evolving of tool sets and frameworks that increasingly live at the nexus of the field of Bioinformatics.
+We've given a general characterization of the contrast that can exist between development schemas that are separated by only 5 years time and the substantive impact to be had in adopting these new paradigms.
 
-Best practices. Discuss open source.
+With that, the CCGD will continue to function as it has since its initial release by offering a point of integration for collated results in the cancer gene screening space.
+Existing connections have been maintained both to and from external resources such as those from the CCGD pointing to the GeneCards suite, NCBI's HomoloGene, and any associated screening publication.
+Additionally, an external gene-level reference link pointing to the CCGD from the Sanger COSMIC database has been maintained as well.
+Finally, new studies will continue to be integrated into the core table as they are published and curated, as was done before.
 
-Mention cosmic connection.
+Finally, we have maintained the essence of that functionality while notably reducing the digital entropy within the codebase and application architecture.
+At the heart of this redesign, we have placed the Rmarkdown web paradigm at the center of the application.
+We feel this offers not only an expanded repertoire of functionality and features but brings the application stack to parity with many of the comparable Bioinformatics web tools of the day.
+In doing so, we implemented our design with an elevated consideration for best practices in development resulting in improved resource outcomes and overall performance.
+This is reflected by qualitative and quantitative comparisons of the codebase.
+We also placed added focus on the nature of the documentation and portability of the code, which is now publicly hosted under the open-source MIT license at github (https://github.com/ctastad/ccgd), (https://doi.org/10.5281/zenodo.4422026).
+With this approach, we hope that the sustained service of the CCGD will continue to serve the research community at-large now and for future refactorizations yet to come.
 

### 2021-05-08 manuscript.tex
performed spellcheck

@@
 Here, we present the case of the Candidate Cancer Gene Database (CCGD, https://z.umn.edu/ccgd) as a minor instance of taking advantage of an upgrade cycle in an attempt to service some of this technical debt.
-The goals of this project ultimately strived to acheive the implementation of a new, core framework at the base of the data integration backend as well as a more modern and simplified web front-end.
+The goals of this project ultimately strived to achieve the implementation of a new, core framework at the base of the data integration backend as well as a more modern and simplified web front-end.
 In all, this represented a complete rebuild of the application while preserving the existing conceptual implementation.
@@
 The markdown output from this step is seamlessly injected into a Pandoc conversion which offers a large variety in document output.
-Specific to our case, the Pandoc conversion paired with a companion YAML configuration file allows for multidocument markdown-to-html rendering which is assembled as a complete site stack.
+Specific to our case, the Pandoc conversion paired with a companion YAML configuration file allows for multi-document markdown-to-html rendering which is assembled as a complete site stack.
 As a single process, this function generates an integrated, modern web front-end from content written in a light-weight syntax with almost none of the typical complexity of full-scale web page creation.}\label{fig:renderFlow}
@@
 Shown are the horizontally oriented file trees of the site directories of the new and old projects.
-What is illustrated is a dramatic improvement in the tree size and complexity through the elimination of a large degree of redudant or unecessary content, log, and style files.
+What is illustrated is a dramatic improvement in the tree size and complexity through the elimination of a large degree of redundant or unnecessary content, log, and style files.
 This is accentuated even further when comparing the source code.

### 2021-05-09 manuscript.tex
wrote description to interface fig; general cleanup

@@
 
-    [full month and year of degree conferral]
+    May, 2021
 
@@
 
-% Acknowledgements
-\newpage
-\pagenumbering{roman}
-
-acknowledgements
-
-% Dedication
-\newpage
-
-dedication
+%% Acknowledgements
+%\newpage
+%\pagenumbering{roman}
+%
+%acknowledgements
+%
+%% Dedication
+%\newpage
+%
+%dedication
 
@@
 Maintenance and upgrades are a fundamental requirement of any software-based tool, and the resources within the field of Bioinformatics are no exception.
-Too often, the combination of using self-taught programmers who are typically trained Biologists along with the broader lack of a software development ecosystem can propagate elevated degrees of software entropy, also known as technical debt.
+Too often, the combination of using self-taught programmers, who are typically trained Biologists, along with the broader lack of a software development ecosystem can propagate elevated degrees of software entropy, also known as technical debt.
 It is then a fair assumption to say that this level of entropy is higher than in fields which more consistently employ the skills of professional developers.
-Here, we present the case of the Candidate Cancer Gene Database (CCGD, https://z.umn.edu/ccgd) as a minor instance of taking advantage of an upgrade cycle in an attempt to service some of this technical debt.
-The goals of this project ultimately strived to achieve the implementation of a new, core framework at the base of the data integration backend as well as a more modern and simplified web front-end.
-In all, this represented a complete rebuild of the application while preserving the existing conceptual implementation.
+Here, we present the case of the Candidate Cancer Gene Database (CCGD, \url{https://z.umn.edu/ccgd}) as a minor instance of taking advantage of an upgrade cycle in an attempt to service some of this technical debt.
+The goals of this project ultimately strived to achieve the implementation of a new, core framework at the base of the data integration backend as well as a more modern and simplified front-end.
+In all, this represented a complete rebuild of the web tool while preserving the existing conceptual implementation.
 Our work served to establish a new application with greater structural simplicity, improved resource use, and, most relevant, a code base which will be easier to maintain and upgrade in the future.
@@
 As a result, a measurable area of focus is not just in the novel discovery of Biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions.
-Further, the crossover of Biological and technical expertise has offered, at times, a cluttered approach to implementing these solutions resulting in less than desirable engineering.
+Further, the crossover of Biological and technical expertise has offered, at times, a cluttered approach to implementing these solutions, resulting in less than desirable engineering.
 
@@
 These improvements have served to enhance the functionality of applications along with creating a reformed development process that is more intuitive, more accessible, and, ultimately, more powerful.
-These advances offer specific utility to the sciences, and following a comparable progress in application bears consideration.
+These advances offer specific utility to the sciences, and, following a comparable progress in application, bears consideration.
 
@@
 This can serve many purposes that all aim to accomplish some range of improvements to the backend design, architecture, or performance.
-Within many engineering goals, this process offers access to items that address routine maintainability to wholesale upgrade.
+Within many engineering goals, this process can range from items that address routine maintainability to wholesale upgrade.
 At times, this may not just be preferable but essential in the scope of scheduled updates and general obsolescence, making it a critical process in the scope of software development.
@@
 This project's goal was ultimately to upgrade the existing CCGD\@.
-In setting out to do this, we established several requirements that took into consideration the opportunity to seize a routine set of server transitions to implement improved functionality.
+In setting out to do this, we established several expanded requirements which seized upon the need to perform a routine set of server transitions, allowing us to implement bolstered functionality.
 Specifically, the central output of the service should be retained in the eyes of the end user.
 Beyond this, attempting to fully reconstruct the backend provided for significant improvements (see Table~\ref{table:projectReqs}).
-These goals more broadly were: upgrade the existing server build to RHEL 7; rewrite web interface in a more modern framework; streamline the central table construction process using Rshiny; implement improved ability to manipulate table and web content by product owner (Tim Starr); employ improved software development best practices; generally seek opportunities for codebase and resource utilization improvements.
+These goals more broadly were: upgrade the existing server build to Red Hat Enterprise Linux 7 (RHEL 7); rewrite the web interface in a more modern framework; streamline the central table construction process using Rshiny; implement an improved ability to manipulate the table and web content by the product owner (Tim Starr); employ improved software development best practices; generally seek opportunities for codebase and resource utilization improvements.
 Ultimately, all of these taken together were intended to serve some form of modernization and simplification.
@@
         \midrule
-        1&server OS&upgrade server OS&RHEL7&Transition of architecture for public server host. This is required by University OIT due to end of life schedule for RHEL6.\\
+        1&server OS&upgrade server OS&RHEL7&Transition of architecture for public server host. This is required by University OIT due to end-of-life schedule for RHEL6.\\
         \midrule
@@
 At it's core, the OS transition from RHEL 6 to 7 was the original project requirement that spurred this work.
-Existing hosting for the CCGD was owned by university OIT who established an end of life date for the current server OS\@.
+Existing hosting for the CCGD was owned by the university Office of Information Technology (OIT) who established an end-of-life date for the current server OS\@.
 As a result, the bulk of core architecture implementation followed a narrow prescription for what was termed a ``Fully Managed'' RHEL7 install as detailed by the OIT-LPT Public-Docs~\cite{oitOITLPTPublicDocsThis}.
@@
 \subsection{Web Framework}
-The upgraded build of the CCGD employed a completely different front-end architecture centralized around R web services environment.
+The upgraded build of the CCGD employed a completely different front-end architecture centralized around the R web services environment.
 Built with dynamic modularity and data portability in mind, the Rmarkdown framework offers the ability to render several independent markdown files into a unified website.
@@
 While following the existing site map, we recreated the raw content of the existing application as independent markdown files.
-Within the Rmarkdown format guide.
-The keystone feature of the Rmarkdown framework that allows for this unified rendering is the use of a YAML configuration header (see Figure~\ref{fig:siteYaml}).
+The keystone feature of the Rmarkdown framework that allows for this unified rendering is the use of a YAML Ain't Markup Language (YAML) configuration header (see Figure~\ref{fig:siteYaml}).
 The \texttt{rmarkdown::render\_site} function leverages this configuration file to align and sequentially render content and styling in tandem for a complete site map [citeme rmarkdown docs].
@@
     \caption[Rmarkdown Rendering Flow]{\textbf{Rmarkdown Rendering Flow} The core web framework was created using a stitch process composed of two document conversion steps, which are both contained in the single \texttt{rmarkdown::render\_site} command.
-These two transformations are done by the R-specific knitr package and then the open-source Pandoc tool.
+These two transformations are done by the R-specific \texttt{knitr} package and then the open-source Pandoc tool.
 The first phase holds less relevance in particular as its role is to execute R code chunks, of which there are few in our page content.
 The markdown output from this step is seamlessly injected into a Pandoc conversion which offers a large variety in document output.
-Specific to our case, the Pandoc conversion paired with a companion YAML configuration file allows for multi-document markdown-to-html rendering which is assembled as a complete site stack.
+Specific to our case, the Pandoc conversion paired with a companion YAML configuration file allows for multi-document markdown-to-html rendering, which is assembled as a complete site stack.
 As a single process, this function generates an integrated, modern web front-end from content written in a light-weight syntax with almost none of the typical complexity of full-scale web page creation.}\label{fig:renderFlow}
@@
 
-An important distinction in summarizing the implementation of web framework is to point out that the Rshiny app, which contains the gene table, is assembled and hosted as a function separate from the general site rendering.
+An important distinction in summarizing the implementation of the web framework is to point out that the Rshiny app, which contains the gene table, is assembled and hosted as a function separate from the general site rendering.
 More specifically to the web design, the shiny app is hosted at the common public repository \href{https://ctastad.shinyapps.io/table_app/}{\texttt{shinyapps.io}} and rendered in an inline iframe on the site search page.
-In the eyes of the user this presents a seamless transition, but the functional contrast permits for differential rendering between the two.
+In the eyes of the user, this presents a seamless transition, but the functional contrast permits for differential rendering between the two.
 This allows for the continuous integration of external sources that feed the table while eliminating the need to regularly render the static site content.
@@
 The central takeaway from this illustration is to recognize the heavy lifting performed by these relatively small code segments.
-Figure~\ref{fig:pageHeader} provides an example of the header found in every~.Rmd file that make-up the respective pages of the site.
+Figure~\ref{fig:pageHeader} provides an example of the header found in every \texttt{.Rmd} file that make-up the respective pages of the site.
 These headers provide primarily formatting configuration for the rendering process but include broader functions as well.
@@
 The most substantial of the changes that addressed these issues was the transition to use of R as the data manipulation toolset.
-More specifically, the~\texttt{build\_table.R} script established a lean extract, transform, load (ETL) process that took the existing table data and external reference data to perform a transformation which collated old and new data.
+More specifically, the~\texttt{build\_table.R} script established a lean extract, transform, load (ETL) process which took the existing table data and external reference data to perform a transformation which collated old and new data.
 The product of this process could then be delivered to the R-based shiny app deployment generated by the~\texttt{app.R} script.
@@
 Redundancy reduction was achieved within script functionality and storage utilization.
-Dedicated script actions that performed house keeping functions, as seen in the~\texttt{build\_table.sh} and~\texttt{backup.sh} (see Figure~\ref{fig:subroutine}), made specific effort to clear all transient data content that otherwise did not require persistent storage.
-This accounted for a large portion of persistent data that was previously left on the server.
+Dedicated script actions that performed house keeping functions, as seen in the~\texttt{build\_table.sh} and~\texttt{backup.sh} scripts (see Figure~\ref{fig:subroutine}), made specific effort to clear all transient data content that otherwise did not require persistent storage.
+This accounted for a large portion of persistent data that was previously left on the server, creating storage constraints.
 Additionally, script reduction was improved through a reorganization of the general subroutines within the application.
 A reassignment of script actions were also implemented in a more cohesive manner which allowed for both improved simplicity and modularity.
-This is most effectively seen in the manual control~\texttt{ccgd\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app (see Figure~\ref{fig:subroutine}) if run manually.
-All of this core functionality was then was automated with scheduled using system cron jobs to sustain routine maintenance in upkeep of the application's table.
+This is most effectively seen in the manual control~\texttt{ccgd\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app if run manually (see Figure~\ref{fig:subroutine}).
+All of this core functionality was then was automated within the system scheduling system using several cron jobs to sustain routine maintenance in upkeep of the application's table.
 
@@
 \subsection{Rshiny App}
-In order to replace the existing CCGD search functionality our group employed the Rshiny framework.
+In order to replace the existing CCGD search functionality, our group employed the Rshiny framework.
 Simply put, the shiny app that now enables this central purpose is an interactive web interface for a single table.
@@
 The broad format of this work flow is \texttt{ui + server $\rightarrow$ deployment}.
-The server specifically has a discrete composition made up of reactive constructs of inputs, expressions, and outputs that shape the app's reactive graph, which forms, more accurately, the discrete operation of the application.
-The assembly and movement of this graph is both emblematic and practical in displaying the modularity and improved resource utilization of the reactive elements.
+The server specifically has a discrete composition made up of reactive constructs of inputs, expressions, and outputs which shape the app's reactive graph.
+This forms, more accurately, the discrete operation of the application.
+The assembly and movement of this graph is both emblematic and practical in displaying the modularity and improved resource utilization of the reactive elements over the previous table paradigm.
 Without belaboring the description of reactive programming, this model employs a declarative programming style that allows for lazy code execution~\cite{wickhamMasteringShinyBook}.
@@
 The reactive web design is intended to allow for the automatic implementation of changed states applied through the alternate paths of the graph.
-This allows for the passive display of a static representation of the table while allowing for dynamic adjustment to filtering inputs through the \texttt{[Var] $\rightarrow$ [Var].values} path.
+This allows for both the passive display of a static representation of the table along with dynamic adjustment to filtering inputs through the \texttt{[Vars] $\rightarrow$ [Vars].values} path.
 Additionally, this declarative model provides for an interactive table array while only drawing computation when the appropriate event is called.}\label{fig:reactiveGraph}
@@
 
-Within the scope of the \href{https://github.com/ctastad/ccgd/blob/master/table_app/app.R}{\texttt{table\_app/app.R}} server function we had a relatively simple goal of defining variable filters across the CCGD data table.
+Within the scope of the \href{https://github.com/ctastad/ccgd/blob/master/table_app/app.R}{\texttt{table\_app/app.R}} server function, we had a relatively simple goal of defining variable filters across the CCGD data table.
 Taking advantage of Rshiny's modularity, we created a two state path for the imported data object that allowed for both a variable-specific filter step along with a static, unchanged display of the table (see Figure~\ref{fig:reactiveGraph}).
@@
 For processing going down the filtered path, independent variable inputs are available to the user which are passed through two expression elements.
-In selecting a variable filter the \texttt{[Vars].values} expression applies a syntactic subsetting of the variable's data list which is then past to the central \texttt{filter.search} where the broader subsetting is carried to the rest of the table.
+In selecting a variable filter the \texttt{[Vars].values} expression applies a syntactic subsetting of the variable's data list which is then passed to the central \texttt{filter.search} function where the broader subsetting is carried to the rest of the table.
 The \texttt{searchTable} function takes the output of the filter and applies several text mutations to add external linking to several associated resources.
@@
 More specifically, we set parameters to create a reactive version of our function which allows for paging, searching, filtering, and sorting in real-time within the shiny server infrastructure [citeme DataTables].
-Through this function alone we established new server paradigm that eliminated the SQL database.
+Through this function alone, we established a new server paradigm that eliminated the SQL database.
 
@@
     \caption[Server-side Processing Flow]{\textbf{Server-side Processing Flow} This diagram illustrates a small but fundamental change applied to the application table processing through the lens of the server-client relationship.
-The ``old'' flow shows the prior table processing procedure, which is highly typical of a classic Linux-Apache-MySQL-PHP (LAMP) stack website.
+The ``old'' flow shows the prior table processing procedure, which is highly typical of a classic LAMP stack website.
 Inputs and calls made by the browser occur from and within a client-side rendering engine which delivers a static request to the server-side SQL database.
@@
 Instead, we leveraged a server-side engine to handle Asynchronous JavaScript and XML (AJAX) requests which are returned with parsed JavaScript Object Notation (JSON), shown in the ``new'' flow.
-This processing flow allows for the use of a single~.csv file as the data source in table presentation, ordering, and pagination, ultimately eliminating the need for a SQL architecture.}\label{fig:parseFlow}
+This processing flow allows for the use of a single \texttt{.csv} file as the data source in table presentation, ordering, and pagination, ultimately eliminating the need for a SQL architecture.}\label{fig:parseFlow}
 \end{figure}
@@
 
-Throughout the course of the project we utilized a private github code repository for hosting and change tracking.
-This work has since been published on a public repository and assigned a DOI number through \href{https://zenodo.org/record/4422027}{Zenodo}.
+Throughout the course of the project, we utilized a private github code repository for hosting and change tracking.
+This work has since been published on a public repository and assigned a DOI number through Zenodo (https://github.com/ctastad/ccgd), (https://doi.org/10.5281/zenodo.4422026).
 Furthering the spirit of project integrity, we expanded on the nature of the documentation for the user and, more prominently, the product owner.
@@
 Continuing with those architecture choices, Rshiny stood as a natural segue from the Rmarkdown front-end to assemble the table application itself.
-By upgrading the web paradigm from that LAMP stack used in the prior version, we were able to shed some of the burden that come with that classic architecture in creating the new table application.
+By upgrading the web paradigm from that LAMP stack used in the prior version, we were able to shed some of the burden that comes with that classic architecture in creating the new table application.
 This was further realized by leveraging the Rshiny toolkit.
@@
 The foundation laid by these architecture changes provided many substantive avenues for improved design.
-Following best practices as well we also developed improved workflows which extended those core structural improvements.
+Following best practices as well, we also developed improved workflows which extended those core structural improvements.
 The most substantive of these gains was found in the complete rebuild performed on the data integration process.
@@
 
+\begin{figure}[H]
+    \center{}
+    \includegraphics[width=0.8\textwidth]{fig/interface.pdf}
+    \caption[Search Interface Changes]{\textbf{Search Interface Changes} Shown above is a screen-shot comparison of the search table interface between the two versions of the application.
+An understated design paradigm was retained for the sake of limiting visual distraction of the content of the table.
+In contrast, the greatest change to the interface can be seen by the fully contained layout of the new design.
+The prior design employed a dedicated information page for every entry within the table, greatly increasing sitemap complexity.
+By making dedicated elements of the table linkable, we were able to replicate the functionality of this distributed layout with a massive improvement in site and navigation simplicity.}\label{fig:interface}
+\end{figure}
+
 \subsection{Resource Use Improvements}
@@
 Notably, several large reference files in the data integration process were being written to disk.
-We assessed these files and identified several candidates for house keeping integration post-process and found ways to compress those that need remain persistent.
+We assessed these files and identified several candidates for house keeping post-process and found ways to compress those that needed to remain persistent.
 This alone generated approximately a 50\% reduction in existing space use.
@@
 It can be difficult to pin point the driving factor in what leads to the accumulation of software entropy, but it is easy to see it exists nonetheless.
-By continuing our emphasis on best practices, we were able to leverage the simplified architecture upgrades alongside elimination of some functional redundancy to generate a working code base that spans across a much more tidy tree.
+By continuing our emphasis on best practices, we were able to leverage simplified architecture upgrades alongside the elimination of some functional redundancy to generate a working code base that spans across a much more tidy tree.
 
@@
 The original to-scale map on the left gives an exploded view distinguishing the updated app treemap in the bottom right corner with all remaining elements of the map belonging to the old version.
-This visual comparison emphasizes the [10x] space decrease made during the upgrade.
+This visual comparison emphasizes the 150x space decrease made during the upgrade.
 The fruits of several design choices are present in this space reduction.
 The most notable impacts of these choices are several massive files eliminated or substantially reduced (noted by asterisk).
-Notably, the dump.sql file shown represented the disproportionate size of the database and space payoff earned by eliminating it.
-Also, it can be seen that the persistent cosmic.txt data file is present in both versions.
+Notably, the \texttt{dump.sql} file shown represented the disproportionate size of the database and space payoff earned by eliminating it.
+Also, it can be seen that the persistent \texttt{cosmic.txt} data file is present in both versions.
 Improvements made in post-process trimming of files like this also produced substantial space returns.
@@
 
-\begin{figure}[H]
-    \center{}
-    \includegraphics[width=0.8\textwidth]{fig/interface.pdf}
-    \caption[Search Interface Changes]{\textbf{Search Interface Changes}}\label{fig:interface}
-\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 This work has given new life to the CCGD and the valuable functionality that it offers in the form of accessible forward genetic cancer screen data.
-Further, it has been a case study in the benefits of evolving of tool sets and frameworks that increasingly live at the nexus of the field of Bioinformatics.
+Further, it has been a case study in the benefits of evolving tool sets and frameworks that increasingly live at the nexus of the field of Bioinformatics.
 We've given a general characterization of the contrast that can exist between development schemas that are separated by only 5 years time and the substantive impact to be had in adopting these new paradigms.
@@
 Additionally, an external gene-level reference link pointing to the CCGD from the Sanger COSMIC database has been maintained as well.
-Finally, new studies will continue to be integrated into the core table as they are published and curated, as was done before.
+Also, new studies will continue to be integrated into the core table as they are published and curated, as was done before.
 
@@
 At the heart of this redesign, we have placed the Rmarkdown web paradigm at the center of the application.
-We feel this offers not only an expanded repertoire of functionality and features but brings the application stack to parity with many of the comparable Bioinformatics web tools of the day.
-In doing so, we implemented our design with an elevated consideration for best practices in development resulting in improved resource outcomes and overall performance.
-This is reflected by qualitative and quantitative comparisons of the codebase.
-We also placed added focus on the nature of the documentation and portability of the code, which is now publicly hosted under the open-source MIT license at github (https://github.com/ctastad/ccgd), (https://doi.org/10.5281/zenodo.4422026).
-With this approach, we hope that the sustained service of the CCGD will continue to serve the research community at-large now and for future refactorizations yet to come.
+We feel this has offered not only an expanded repertoire of functionality and features but has brought the application stack to parity with many of the comparable Bioinformatics web tools of the day.
+In doing so, we implemented our design with an elevated consideration for best practices in development, resulting in improved performance and overall codebase health.
+This is reflected by qualitative and quantitative comparisons of the resource usage of the application.
+We also placed added focus on the nature of the documentation and portability of the code, which is now publicly hosted at github under the open-source MIT license.
+With this approach, we hope that the sustained service of the CCGD will continue to serve the research community at-large, now and for future refactorizations yet to come.
 

### 2021-05-09 manuscript.tex
final polishing

@@
 It is then a fair assumption to say that this level of entropy is higher than in fields which more consistently employ the skills of professional developers.
-Here, we present the case of the Candidate Cancer Gene Database (CCGD, \url{https://z.umn.edu/ccgd}) as a minor instance of taking advantage of an upgrade cycle in an attempt to service some of this technical debt.
+Here, we present the case of the Candidate Cancer Gene Database (CCGD, \url{https://z.umn.edu/ccgd}), a web application offering accessible forward transposon screen results in mice, as a minor instance of taking advantage of an upgrade cycle in an attempt to service some of this technical debt.
 The goals of this project ultimately strived to achieve the implementation of a new, core framework at the base of the data integration backend as well as a more modern and simplified front-end.
@@
 \chapter{Introduction}
-In many ways, the field of Bioinformatics exists as the result of a problem rather than a discovery, as with other scientific fields.
-The nature of the challenges have evolved from a growth in the composition of primary data and a lacking compensatory set of technical skills to evaluate it.
+In many ways, the field of Bioinformatics exists as the result of a set of challenges rather than a discovery, as with other scientific fields.
+The nature of the challenges have evolved from a growth in the composition of primary data and a lacking compensatory set of technical skills to evaluate it~\cite{lawlorEngineeringBioinformaticsBuilding2015}.
 With the bottle neck that exists at the center of this problem, previously simple notions of collating, analyzing, and visualizing results no longer scale.
-As a result, a measurable area of focus is not just in the novel discovery of Biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions.
-Further, the crossover of Biological and technical expertise has offered, at times, a cluttered approach to implementing these solutions, resulting in less than desirable engineering.
+As a result, a measurable area of focus is not just in the novel discovery of Biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions~\cite{segalDevelopingScientificSoftware2008}.
+Further, the crossover of Biological and technical expertise has offered, at times, a cluttered approach to implementing these solutions, resulting in less than desirable engineering~\cite{vermaLackSoftwareEngineering2013}.
 
-All while this dynamic has developed in the world of Biology, the technology has rapidly improved.
-A pillar of the field's advances is in the area of web technologies, which now offer frameworks and tools that both accomplish more while requiring less of their developers.
-The classical paradigm of the web has evolved from the LAMP stack (Linux, Apache, MySQL, PHP) to improved frameworks such as advanced JavaScript, CSS3, and HTML5 that lay the foundation for modern and flexible web applications.
+All while this dynamic has developed in the world of Biology, the technology has rapidly improved~\cite{teufelCurrentBioinformaticsTools2006}.
+A pillar of the field's advances is in the area of web technologies, which now offer frameworks and tools that both accomplish more while requiring less of their developers~\cite{managementassociationBioinformaticsConceptsMethodologies2013a}.
+The classical paradigm of the web has evolved from the LAMP stack (Linux, Apache, MySQL, PHP) to improved frameworks such as advanced JavaScript, CSS3, and HTML5 that lay the foundation for modern and flexible web applications~\cite{wangOpenSourceLibraries2015}.
 These improvements have served to enhance the functionality of applications along with creating a reformed development process that is more intuitive, more accessible, and, ultimately, more powerful.
-These advances offer specific utility to the sciences, and, following a comparable progress in application, bears consideration.
+These advances offer specific utility to the sciences and bears consideration.
 
-While the implementation of new models typically applies to future work, it is essential to look back at opportunities to improve past engineering with updated models through the process of code refactoring.
-At a basic level, refactoring is simply restructuring or rewriting the content of a codebase while retaining the functionality of that service.
+While the implementation of new models typically applies to future work, it is essential to look back at opportunities to improve past engineering with updated models through the process of code refactoring~\cite{fowlerRefactoringImprovingDesign1999}.
+At a basic level, refactoring is simply restructuring or rewriting the content of a codebase while retaining the functionality of that service~\cite{kerievskyRefactoringPatterns2004}.
 This can serve many purposes that all aim to accomplish some range of improvements to the backend design, architecture, or performance.
 Within many engineering goals, this process can range from items that address routine maintainability to wholesale upgrade.
-At times, this may not just be preferable but essential in the scope of scheduled updates and general obsolescence, making it a critical process in the scope of software development.
+At times, this may not just be preferable but essential in the scope of scheduled updates and general obsolescence, making it a critical process in the scope of software development~\cite{kerievskyRefactoringPatterns2004}.
 
@@
 With these incremental changeovers, no mater the impact, the field takes on a new character in the management of its growing data challenges.
-The CCGD as a concept set out to tackle some of the original problems of filtering handling large quantities of data.
-We expand on this by re-implementing the same achievement through a modern application.
-
-%Gene screens in cancer bio~\cite{Clark2019}
-
-%Cancer gene screening is a hallmark case for this exploration as it lives at the nexus of being an area of high impact research and a process that can produce a deluge of raw data.
-In tandem, the state of web front ends has advanced in a manner that more completely allows~\cite{10.1093/nar/gkz1161}
+The CCGD as a concept set out to tackle some of the original problems of filtering large quantities of data.
+More specifically, the CCGD exists as a repository for the results of forward genetic screens in transposon mutagenesis in mice~\cite{abbottCandidateCancerGene2015a}.
+The nature of cancer biology offers a highly complex genetic landscape with numerous phenotypes.
+While many investigators start with goals to identify more focused targets in the spirit of achieving precision therapies, an essential step in this process begins with whole-sale filtering of candidates for cancer driver genes~\cite{clarkCancerGeneDiscovery2019}.
+Dramatic impacts have been made in the field by the identification of important cancer genes, and the addition of items like transposon screen results quality scores and strength metrics adds to the ability to make this progress.
+We then expand on this concept by re-implementing the same achievement within a modern framework.
 
@@
 \subsection{Web Framework}
-The upgraded build of the CCGD employed a completely different front-end architecture centralized around the R web services environment.
-Built with dynamic modularity and data portability in mind, the Rmarkdown framework offers the ability to render several independent markdown files into a unified website.
+The upgraded build of the CCGD employed a completely different front-end architecture centralized around the R web services environment~\cite{rstudioteamRStudioIntegratedDevelopment2020}.
+Built with dynamic modularity and data portability in mind, the Rmarkdown framework offers the ability to render several independent markdown files into a unified website~\cite{xieMarkdownDefinitiveGuide}.
 We applied this feature set to produce our general site structure and content as it allowed for a massive simplification in web design.
@@
 The keystone feature of the Rmarkdown framework that allows for this unified rendering is the use of a YAML Ain't Markup Language (YAML) configuration header (see Figure~\ref{fig:siteYaml}).
-The \texttt{rmarkdown::render\_site} function leverages this configuration file to align and sequentially render content and styling in tandem for a complete site map [citeme rmarkdown docs].
+The \texttt{rmarkdown::render\_site} function leverages this configuration file to align and sequentially render content and styling in tandem for a complete site map~\cite{xieMarkdownDefinitiveGuide}.
 Additionally, individual markdown pages employ this same configuration paradigm for page-specific parameters and formatting guidelines (see Figure~\ref{fig:pageHeader}).
@@
     \caption[Rmarkdown Rendering Flow]{\textbf{Rmarkdown Rendering Flow} The core web framework was created using a stitch process composed of two document conversion steps, which are both contained in the single \texttt{rmarkdown::render\_site} command.
-These two transformations are done by the R-specific \texttt{knitr} package and then the open-source Pandoc tool.
+These two transformations are done by the R-specific knitr package and then the open-source Pandoc tool.
 The first phase holds less relevance in particular as its role is to execute R code chunks, of which there are few in our page content.
@@
 
-responsive web design
-
 An important distinction in summarizing the implementation of the web framework is to point out that the Rshiny app, which contains the gene table, is assembled and hosted as a function separate from the general site rendering.
-More specifically to the web design, the shiny app is hosted at the common public repository \href{https://ctastad.shinyapps.io/table_app/}{\texttt{shinyapps.io}} and rendered in an inline iframe on the site search page.
+More specifically to the web design, the shiny app is hosted at the common public repository \texttt{shinyapps.io} (\url{https://ctastad.shinyapps.io/table_app/}) and rendered in an inline iframe on the site search page.
 In the eyes of the user, this presents a seamless transition, but the functional contrast permits for differential rendering between the two.
@@
         \includegraphics[width=\textwidth]{fig/page_header.pdf}
-        \subcaption{index.Rmd individual page header}\label{fig:pageHeader}
+        \subcaption{\texttt{index.Rmd} individual page header}\label{fig:pageHeader}
     \end{subfigure}%
@@
         \includegraphics[width=\textwidth]{fig/site_header.pdf}
-        \caption{\_site.yml: file for site assembly}\label{fig:siteYaml}
+        \caption{\texttt{\_site.yml} file for site assembly}\label{fig:siteYaml}
     \end{subfigure}
@@
 Either way, a few lines of configuration generate wildly different outputs.
-Figure~\ref{fig:siteYaml} presents the YAML file which serves to assemble independent Rmarkdown files into a joined website as described in figure~\ref{fig:renderFlow}.
+Figure~\ref{fig:siteYaml} presents the YAML file which serves to assemble independent Rmarkdown files into a joined website as described in Figure~\ref{fig:renderFlow}.
 There is less configuration here as the only real purpose is to define the site tree for the \texttt{rmarkdown::render\_site} process.}\label{fig:yamls}
@@
 The prior implementation employed some tools that are common place, but they were implemented in a fashion that lacked simplicity at times.
-Also, certain proprietary widgets, such as the need for conversion in and out of the Microsoft Excel format, presented functionality and compatibility choke points that generated road blocks to maintenance.
+Also, certain proprietary widgets, such as the need for conversion in and out of the Microsoft Excel format, presented functionality and compatibility choke points that generated road blocks to maintenance~\cite{abbottCandidateCancerGene2015a}.
 The most substantial of the changes that addressed these issues was the transition to use of R as the data manipulation toolset.
@@
 Additionally, script reduction was improved through a reorganization of the general subroutines within the application.
-A reassignment of script actions were also implemented in a more cohesive manner which allowed for both improved simplicity and modularity.
+A reassignment of script actions was also implemented in a more cohesive manner which allowed for both improved simplicity and modularity.
 This is most effectively seen in the manual control~\texttt{ccgd\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app if run manually (see Figure~\ref{fig:subroutine}).
-All of this core functionality was then was automated within the system scheduling system using several cron jobs to sustain routine maintenance in upkeep of the application's table.
+All of this core functionality was then was automated within system scheduling using several cron jobs to sustain routine maintenance in upkeep of the application's table.
 
@@
 In order to replace the existing CCGD search functionality, our group employed the Rshiny framework.
-Simply put, the shiny app that now enables this central purpose is an interactive web interface for a single table.
+Simply put, the shiny app that now enables this central purpose is an interactive web interface for a single table~\cite{changShinyWebApplication2021}.
 Several changes were made to the existing integration pipeline in order to achieve this simplicity.
@@
 The server specifically has a discrete composition made up of reactive constructs of inputs, expressions, and outputs which shape the app's reactive graph.
-This forms, more accurately, the discrete operation of the application.
+This forms, more accurately, the discrete operation of the application~\cite{wickhamMasteringShinyBook}.
 The assembly and movement of this graph is both emblematic and practical in displaying the modularity and improved resource utilization of the reactive elements over the previous table paradigm.
-Without belaboring the description of reactive programming, this model employs a declarative programming style that allows for lazy code execution~\cite{wickhamMasteringShinyBook}.
+Without belaboring the description of reactive programming, this model employs a declarative programming style that allows for lazy code execution.
 What this amounts to in application is for the developer to define loose controls at a higher level while allowing the code to passively fill these requirements.
-This translates to greater flexibility to both the developer and the machine as resource use can be reduced through modularity in code execution.
+This translates to greater flexibility to both the developer and the machine as resource use can be reduced through modularity in code execution~\cite{jwlloydPracticalAdvantagesDeclarative1994}.
 
@@
 The prior iteration of the CCGD architecture employed a relatively disproportionate Structured Query Language (SQL) database to store and serve numerous component elements that made up the table.
-Instead, the new implementation of our app now leverages the server-side processing capability of the DataTables JavaScript library.
-More specifically, we set parameters to create a reactive version of our function which allows for paging, searching, filtering, and sorting in real-time within the shiny server infrastructure [citeme DataTables].
+Instead, the new implementation of our app now leverages the server-side processing capability of the DataTables JavaScript library~\cite{DataTablesTablePlugin}.
+More specifically, we set parameters to create a reactive version of our function which allows for paging, searching, filtering, and sorting in real-time within the shiny server infrastructure.
 Through this function alone, we established a new server paradigm that eliminated the SQL database.
@@
 
-In approaching the process of development, we applied a philosophy of tidy code creation to both syntax and design.
+In approaching the process of development, we applied a philosophy of tidy code creation to both syntax and design~\cite{wickhamDplyrGrammarData2021}.
 A supporting method in creating a more elegant code syntax was the use of R universe tools, Tidyverse and Rmarkdown.
@@
 Following the goals set out in our project requirements, we chose Rmarkdown as a replacement framework to develop the application's web front-end.
-Previously, the web interface for the application had been developed using Adobe Dreamweaver~\cite{abbottCandidateCancerGene2015a} resting on a traditional LAMP stack.
+Previously, the web interface for the application had been developed using Adobe Dreamweaver resting on a traditional LAMP stack~\cite{abbottCandidateCancerGene2015a}.
 For the time, a LAMP stack and Dreamweaver served the purpose of simplifying the process of interface construction on a stable and consistent web platform, yet we hoped to extend the theme of simplified content creation into a modern framework.
@@
 Through the Rmarkdown toolset, the work of web design is now largely rendered automatically for the developer (see Figure~\ref{fig:renderFlow}).
-Employing the R web scheme established a cornerstone for the project to leverage both enhanced web development as well as making available other key tools in architecture design choices.
+Employing the R web schema established a cornerstone for the project to leverage both enhanced web development as well as making available other key tools in architecture design choices.
 
@@
 
-Along with software design changes, there were several improvements made to the interface and site appearance.
+Along with software design changes, there were several improvements made to the interface and site appearance (see Figure~\ref{fig:interface}).
 The upgraded web front-end not only offered a needed facelift to the aesthetic appearance of the site, but the new framework brought many of the elements of the page construction current with existing browser standards.
@@
 Minor usability factors were considered as well such as the creation of an improved URL\@.
-The CCGD can now be accessed via~\href{https://z.umn.edu/ccgd}{(z.umn.edu/ccgd)} rather than the much more unruly~\href{http://hst-ccgd-prd-web.oit.umn.edu/}{(http://hst-ccgd-prd-web.oit.umn.edu/)}.
+The CCGD can now be accessed via a shortened URL~\href{https://z.umn.edu/ccgd}{(z.umn.edu/ccgd)} rather than the much more unruly OIT-hosted URL~\href{http://hst-ccgd-prd-web.oit.umn.edu/}{(http://hst-ccgd-prd-web.oit.umn.edu/)}.
 A more major usability achievement was derived from the structure of the new Rshiny table interface.
 With the combination of the reorganized data integration process, the elimination of the SQL backend, and the front-end chosen in the DataTables library, we were able to assemble the application as a single page.
-This dramatically reduced broader site map complexity and simplified the process of interacting with the same data set for the user overall.
+This dramatically reduced broader site map complexity and simplified the process of interacting with the same data set for the user overall (see Figure~\ref{fig:interface}).
 
@@
 
-Another broad concept we addressed in elevating general efficiency was exploring general concepts of technical debt.
+Another concept we addressed in elevating general efficiency was exploring broader concepts of technical debt.
 This idea is addressed in our approach to dealing with space constraints on some level, but it is a problem that can generally permeate an entire codebase.
@@
 \end{figure}
-
-Ref crossdirstat
 
@@
 At the heart of this redesign, we have placed the Rmarkdown web paradigm at the center of the application.
-We feel this has offered not only an expanded repertoire of functionality and features but has brought the application stack to parity with many of the comparable Bioinformatics web tools of the day.
+We feel this has offered not only an expanded repertoire of functionality and features but has brought the application stack to parity with many of the comparable Bioinformatics web tools of the day~\cite{zhengComprehensiveReviewWeb2020a}.
 In doing so, we implemented our design with an elevated consideration for best practices in development, resulting in improved performance and overall codebase health.
@@
 \addcontentsline{toc}{chapter}{Bibliography}
+\nocite{*}
 \printbibliography{}

### 2021-05-20 presentation.Rmd
converted presentation

---
title: "Cool Title"
author: "Chris Tastad"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
  revealjs::revealjs_presentation:
    center: false
    theme: night
    highlight: zenburn
    transition: none
    reveal_options:
      controls: false
      height: "100%"
      margin: 0.04
      slideNumber: true
      previewLinks: true
---

## Technical Debt fig

## CCGD Overview {.center}

## Breakdown of build table process

```{r global_options, include = F}
knitr::opts_chunk$set(eval = F, echo = F, warning = F, message = F)
```

## Application Concept

## Forward Transposon Screens

## Project Requirements

::: {#table:projectReqs}
      Feature           Goal                       Framework   Description
  --- ----------------- -------------------------- ----------- -------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  1   server OS         upgrade server OS          RHEL7       Transition of architecture for public server host. This is required by University OIT due to end-of-life schedule for RHEL6.
  2   web front-end     rewrite web interface      Rmarkdown   Improvements to the web interface written in a modern, simplified language. This improved access to content creation and allowed for automation in front-end rendering.
  3   table build       rewrite table build        Rshiny      Rshiny offered a dramatic improvement to replace the existing process by merging the table build back-end with a modern web display of the app interface.
  4   content update    improved admin controls    BASH/R      Old app version confined some content controls to app author, limiting ability to make contributions by product owner (Tim Starr).
  5   version control   implement best practices   git/docs    No version control was used in the original development of the app. This and other documentation practices were expanded in the rewrite.
  6   multiple          resource improvements      codebase    Due to the lack of some best practices, there were many opportunities to make impactful resource improvements.

:::

## Rmarkdown Rendering Flow
<img src="./fig/rmarkdown_flow.svg">

## YAML Configuration Segments

<div class="column">

```{r echo = T}
title: "Candidate Cancer Gene Database"
bibliography: refs/ccgd_paper.bib
nocite: '@*'
output:
    html_document:
        includes:
            in_header: "styles/favicon.html"
            after_body: "styles/footer.html"
        css: styles/styles.css
        theme: readable
---

```

```{r echo = T}
name: "Candidate Cancer Gene Database"
navbar:
    title: "CCGD"
    left:
        - text: "Home"
          href: index.html
        - text: "Search"
          href: search.html
        - text: "Help"
          href: help.html
        - text: "References"
          href: references.html
        - text: "Contact"
          href: contact.html

```

</div>

![](./fig/yaml_illustration.svg)

## Subroutine Flow
![](./fig/subroutine_flow.svg)

## Table Reactive Graph
![](./fig/reactive_graph.svg)

## Server-side Processing Flow
![](./fig/parse_flow.svg)

## Search Interface Changes
![](./fig/interface.svg)

## Disk Use Comparison
![](./fig/space_tree.svg)

## Project Directory File Tree Comparison
![](./fig/file_tree.svg)

## Documentation

## app.R code

## markdown code

## List of app improvements

## The End {.center}

<!-- style -->

<style>

body {
  background: #2e3440;
}

.column {
  column-count: 2;
}

div.sourceCode {
  background-color: #2e3440;
}

.sourceCode {
  background-color: #393f4b;
}

.reveal .progress span {
  background: #7f9f7f;
}

.reveal section img{
  border: 30px solid #e0e1e2;
  border-radius: 10px;
  }

</style>

<!-- slide templates

## Section Title {.center}

## Multi-Column Slide

<div class="column">

### column1
- some cool words
- https://www.merriam-webster.com/dictionary/wherever

### column2
- some more cool words

</div>

## Figure Slide
```{r eval=T}
attach(mtcars)
par(mfrow=c(2,2))
plot(wt,mpg, main="Scatterplot of wt vs. mpg")
plot(wt,disp, main="Scatterplot of wt vs disp")
hist(wt, main="Histogram of wt")
boxplot(wt, main="Boxplot of wt")
```

## Code Slide

```{r echo = T}
x = 1
if x == 1:
    # indented four spaces
    print("x is 1.")
```

-->

### 2021-05-20 presentation.Rmd
added a few more figs; expanded presentation content

@@
 ---
-title: "Cool Title"
-author: "Chris Tastad"
-date: "`r format(Sys.time(), '%b %d, %Y')`"
+title: "Refactoring the Candidate Cancer Gene Database"
+author: "Christopher Tastad"
+date: "May 21, 2021"
 output:
   revealjs::revealjs_presentation:
+    self_contained: false
     center: false
@@
     transition: none
+    reveal_plugins: ["menu", "notes"]
     reveal_options:
-      controls: false
+      showNotes: "separate-page"
+      controls: true
       height: "100%"
@@
 
-## Technical Debt fig
-
-## CCGD Overview {.center}
-
-## Breakdown of build table process
+## Presentation ToC
+> - Original Concept
+> - Challenges
+> - Refactoring the Concept
+> - Outcomes of Refactoring
+> - Conclusions
 
@@
 
-## Application Concept
+# Original Concept {.center}
+
+--------------------------------------------------------------------------------
 
 ## Forward Transposon Screens
+
+--------------------------------------------------------------------------------
+
+## CCGD Overview
+
+--------------------------------------------------------------------------------
+
+## Breakdown of build table process
+
+--------------------------------------------------------------------------------
+
+## OIT setup
+
+################################################################################
+
+# Challenges {.center}
+
+--------------------------------------------------------------------------------
+
+## Forced Upgrades
+
+--------------------------------------------------------------------------------
+
+## Technical Debt
+> - test
+> - test
+
+################################################################################
+
+# Refactoring the Concept {.center}
+
+--------------------------------------------------------------------------------
 
@@
 
-## Rmarkdown Rendering Flow
-<img src="./fig/rmarkdown_flow.svg">
+--------------------------------------------------------------------------------
 
-## YAML Configuration Segments
+## Front-end Framework
+![](./fig/rmarkdown_flow.svg)
+
+#### Rmarkdown rendering flow {.center}
+
+--------------------------------------------------------------------------------
+
+## Front-end Framework
 
@@
 
-![](./fig/yaml_illustration.svg)
+#### YAML configuration segments {.center}
 
-## Subroutine Flow
+<center>
+<img src="./fig/yaml_illustration.svg">
+</center>
+
+#### YAML illustration
+
+--------------------------------------------------------------------------------
+
+## Front-end Framework
+![](./fig/reactive_graph.svg)
+
+#### table reactive graph {.center}
+
+--------------------------------------------------------------------------------
+
+## Back-end Framework
+![](./fig/parse_flow.svg)
+
+#### server-side processing flow {.center}
+
+--------------------------------------------------------------------------------
+
+## Back-end Framework
 ![](./fig/subroutine_flow.svg)
 
-## Table Reactive Graph
-![](./fig/reactive_graph.svg)
+#### subroutine flow {.center}
 
-## Server-side Processing Flow
-![](./fig/parse_flow.svg)
+################################################################################
 
-## Search Interface Changes
+# Outcomes of Refactoring {.center}
+
+--------------------------------------------------------------------------------
+
+## Simplified Codebase
+```{r echo = T}
+df <- read.csv("ccgd_export.csv") # read in base source file
+speciesList <- c("Mouse", "Human", "Rat", "Fly", "Fish", "Yeast")
+
+# inputs are arranged in column, width orientation
+# each column variable set represents a single input and its params
+# next row of inputs layout
+ui <- fluidPage( # build shiny app UI
+  fluidRow(
+    column(
+      2,
+      selectizeInput("Species",
+        label = "Species",
+        choices = speciesList
+      )
+    ),
+
+    column(
+      3,
+      selectizeInput("Study",
+        label = "Study",
+        choices = sort(unique(df$Study)),
+        selected = NULL,
+        multiple = TRUE,
+        options = list(placeholder = "All studies")
+      )
+    ),
+
+    column(
+      3,
+      selectizeInput("Cancer",
+        label = "Cancer",
+        choices = sort(unique(df$Cancer)),
+        selected = NULL,
+        multiple = TRUE,
+        options = list(placeholder = "All cancers")
+      )
+    ),
+
+    column(
+      4,
+      textAreaInput("Genes",
+        label = "Genes",
+        placeholder = "GeneA,GeneB,GeneC..."
+      )
+    )
+  ),
+
+  fluidRow(
+    column(
+      2,
+      # button for table export
+      downloadButton("downloadData",
+        label = "Download"
+      )
+    )
+  ),
+
+  hr(),
+
+  dataTableOutput("searchTable")
+)
+```
+
+#### app.R Rshiny process snippet {.center}
+
+--------------------------------------------------------------------------------
+
+## Simplified Codebase
+
+```{r echo = T}
+---
+title: "Candidate Cancer Gene Database"
+bibliography: refs/ccgd_paper.bib
+nocite: '@*'
+output:
+    html_document:
+        includes:
+            in_header: "styles/favicon.html"
+            after_body: "styles/footer.html"
+        css: styles/styles.css
+        theme: readable
+---
+
+The Candidate Cancer Gene Database (CCGD) was developed to make accessible a collated set of results from transposon-based forward cancer genetic screens in mice.
+The value in doing this is to give investigators the ability to quickly filter through the results of many such screens in an effort to determine the candidacy of a gene for its role in cancer.
+
+***
+
+### Background
+
+This work is a product of the [Starr Lab](http://www.med.umn.edu/starrlab/) at the [University of Minnesota](http://www.umn.edu/).
+The Starr Lab is affiliated with the [Center for Genome Engineering](http://www.cge.umn.edu/), the [Masonic Cancer Center](http://www.cancer.umn.edu/), the Department of [OB/GYN](http://www.obgyn.umn.edu/), and the Department of [Genetics, Cell Biology, and Development](http://www.cbs.umn.edu/gcd/).
+
+##### Contributions
+
+The idea for creating this database originated from Tim Starr, Vincent Keng, and David Largaespada at UMN.
+Erik Nyre and Ken Abbott developed the original database, with help from Juan Abrahante.
+A complete website and backend overhaul was conducted by Christopher Tastad in 2019.
+
+* Ken Abbott
+* [Erik Nyre](https://www.dorsey.com/people/n/nyre-erik)
+* [Juan Abrahante](https://research.umn.edu/units/umii/about-us/staff-directory/juan-e-abrahante-phd)
+* [Christopher Tastad](https://github.com/ctastad)
+
+***
+
+##### Updates
+
+The most recent table build was completed on:
+
+<object type="text/html"
+        height=40px
+        data="build_date.txt"></object>
+
+```
+
+#### index.html page content snippet {.center}
+
+--------------------------------------------------------------------------------
+
+## Simplified Codebase
+![](./fig/file_tree.svg)
+
+#### Project Directory File Tree Comparison {.center}
+
+--------------------------------------------------------------------------------
+
+## Visual Improvements
 ![](./fig/interface.svg)
 
-## Disk Use Comparison
+#### search interface changes {.center}
+
+--------------------------------------------------------------------------------
+
+## Resource Use Reduction
 ![](./fig/space_tree.svg)
 
-## Project Directory File Tree Comparison
-![](./fig/file_tree.svg)
+#### disk use comparison {.center}
 
-## Documentation
+--------------------------------------------------------------------------------
 
-## app.R code
+## Improved Admin Controls
 
-## markdown code
+```{r echo = T}
+#   Executed:   locally from the CCGD project dir
+#   Options:    -b build (TRUE,FALSE) -s full proj dir sync (TRUE,FALSE)
+#               -t table upload file -r reference upload file
+#               -k (TURE,FALSE) render site files
 
-## List of app improvements
+# pass arguments from cli
+while getopts b:t:r:s: option
+do
+    case "${option}"
+        in
+        b) build=${OPTARG};;
+        t) table=${OPTARG};;
+        r) refs=${OPTARG};;
+        s) sync=${OPTARG};;
+    esac
+done
+```
+
+#### ccgd_upload.sh script snippet {.center}
+
+--------------------------------------------------------------------------------
+
+## Sustainable Development
+![](./fig/documentation.png)
+
+#### project documentation example {.center}
+
+--------------------------------------------------------------------------------
+
+## Sustainable Development
+
+![](./fig/github.png)
+
+#### public application github repository
+
+--------------------------------------------------------------------------------
+
+## Sustainable Development
+
+![](./fig/release.png)
+
+#### Zenodo-supplied DOI and release for reproducibility {.center}
+
+################################################################################
+
+# Conclusions {.center}
+
+## Conclusions
+
+## Acknowledgements
 
@@
 
-.reveal .progress span {
-  background: #7f9f7f;
+.reveal section img {
+  border: 30px solid #e0e1e2;
+  border-radius: 10px;
   }
 
-.reveal section img{
-  border: 30px solid #e0e1e2;
-  border-radius: 10px;
+  .reveal p {
+    text-align: left;
+  }
+  .reveal ul {
+    display: block;
+  }
+  .reveal ol {
+    display: block;
   }

### 2021-05-20 pres_knit.R

@@ -1,0 +1,4 @@
+#!/usr/bin/env Rscript
+
+rmarkdown::render("presentation.Rmd")
+

### 2021-05-20 manuscript.tex
implemented comments per tim

@@
 With the bottle neck that exists at the center of this problem, previously simple notions of collating, analyzing, and visualizing results no longer scale.
-As a result, a measurable area of focus is not just in the novel discovery of Biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions~\cite{segalDevelopingScientificSoftware2008}.
-Further, the crossover of Biological and technical expertise has offered, at times, a cluttered approach to implementing these solutions, resulting in less than desirable engineering~\cite{vermaLackSoftwareEngineering2013}.
+As a result, a measurable area of focus is not just in the novel discovery of biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions~\cite{segalDevelopingScientificSoftware2008}.
+Further, the crossover of biological and technical expertise has offered, at times, a cluttered approach to implementing these solutions, resulting in less than desirable engineering~\cite{vermaLackSoftwareEngineering2013}.
 
@@
 The nature of cancer biology offers a highly complex genetic landscape with numerous phenotypes.
-While many investigators start with goals to identify more focused targets in the spirit of achieving precision therapies, an essential step in this process begins with whole-sale filtering of candidates for cancer driver genes~\cite{clarkCancerGeneDiscovery2019}.
+While many investigators start with goals to identify more focused targets in the spirit of achieving precision therapies, an essential step in this process begins with wholesale filtering of candidates for cancer driver genes~\cite{clarkCancerGeneDiscovery2019}.
 Dramatic impacts have been made in the field by the identification of important cancer genes, and the addition of items like transposon screen results quality scores and strength metrics adds to the ability to make this progress.
@@
 The header can be as simple as listing a title and Pandoc output type or include LaTex bibliography functionality and read-in style sheets like the one shown.
-Either way, a few lines of configuration generate wildly different outputs.
+Either way, a few lines of configuration generate significantly different outputs.
 Figure~\ref{fig:siteYaml} presents the YAML file which serves to assemble independent Rmarkdown files into a joined website as described in Figure~\ref{fig:renderFlow}.
@@
 Improvements to the data integration process were accomplished using a more simplified framework to eliminate redundancy and provide a more standardized schema.
-The prior implementation employed some tools that are common place, but they were implemented in a fashion that lacked simplicity at times.
+The prior implementation employed some tools that are commonplace, but they were implemented in a fashion that lacked simplicity at times.
 Also, certain proprietary widgets, such as the need for conversion in and out of the Microsoft Excel format, presented functionality and compatibility choke points that generated road blocks to maintenance~\cite{abbottCandidateCancerGene2015a}.
@@
 Redundancy reduction was achieved within script functionality and storage utilization.
-Dedicated script actions that performed house keeping functions, as seen in the~\texttt{build\_table.sh} and~\texttt{backup.sh} scripts (see Figure~\ref{fig:subroutine}), made specific effort to clear all transient data content that otherwise did not require persistent storage.
+Dedicated script actions that performed housekeeping functions, as seen in the~\texttt{build\_table.sh} and~\texttt{backup.sh} scripts (see Figure~\ref{fig:subroutine}), made specific effort to clear all transient data content that otherwise did not require persistent storage.
 This accounted for a large portion of persistent data that was previously left on the server, creating storage constraints.
@@
 The server specifically has a discrete composition made up of reactive constructs of inputs, expressions, and outputs which shape the app's reactive graph.
-This forms, more accurately, the discrete operation of the application~\cite{wickhamMasteringShinyBook}.
+These elements form the discrete operation of the application~\cite{wickhamMasteringShinyBook}.
 The assembly and movement of this graph is both emblematic and practical in displaying the modularity and improved resource utilization of the reactive elements over the previous table paradigm.

### 2021-05-21 presentation.Rmd
furthered presentation content

@@
       margin: 0.04
-      slideNumber: true
+      slideNumber: "c/t"
       previewLinks: true
@@
 > - Original Concept
-> - Challenges
+> - Issue
 > - Refactoring the Concept
@@
 
+## Original Concept
+
+![](./fig/ccgd_header.png)
+
+> - Cancer bio is complicated.
+> - Driver genes present primary targets.
+> - Screening methods...
+
+<div class="notes">
+- look at the origins and maintenance of the ccgd
+- why does such a tool exist
+- what purpose does it offer the field
+
+- Cancer biology presents a highly complex genetic landscape.
+- Cohorts of genes act as drivers for the progression of cancer.
+- These genes present targets for study and development of therapeutics.
+- Ultimately, we need to understand and simplify cancer gene targets.
+- Screening methods offer quicker access to wide swaths of these targets.
+</div>
+
 --------------------------------------------------------------------------------
@@
 
---------------------------------------------------------------------------------
+![](./fig/transposon_screen_mech.jpg)
 
-## CCGD Overview
+<small>Ranzani, Marco, et al. “Cancer Gene Discovery: Exploiting Insertional Mutagenesis.” Mol Cancer Res, vol. 11, no. 10, 2013, pp. 1141–1158., doi:10.1158/1541-7786.MCR-13-0244.</small>
+
+<div class="notes">
+- gene cassette is used to place a genetic payload at some region relative to a transcriptional site to induce mutagenesis
+- figure shows 3 mechanisms to derive outcome
+- A: enhancer insertion likely upstream of promoter.
+- B: vector integrates within some transcriptional range and co-ops promotor function from gene itself.
+- C: intragenic insertions may disrupt coding regions generating novel proteins with GoF or LoF outcomes impacting tumorigenesis or tumor suppression.
+</div>
 
@@
 
-## Breakdown of build table process
+## Forward Transposon Screens
+
+![](./fig/transposon_screen.jpg)
+
+<small>Ranzani, Marco, et al. “Cancer Gene Discovery: Exploiting Insertional Mutagenesis.” Mol Cancer Res, vol. 11, no. 10, 2013, pp. 1141–1158., doi:10.1158/1541-7786.MCR-13-0244.</small>
+
+<div class="notes">
+- transposon method is applied en masse within mice.
+- process of extraction, isolation, seq, and analysis identifies CIS groups.
+- validation takes place in vitro or in vivo for oncogenic potential with orthologous association to human candidates.
+- idea for CCGD comes in at the star.
+</div>
 
@@
 
-## OIT setup
+## CCGD Goals
+
+> To facilitate analysis of driver genes to catalog all common insertion sites (CISs) and their corresponding genes identified in published studies using transposon insertional mutagenesis.
+
+> - Create an interactive web tool to quickly refine the stated data set by a variety of parameters.
+> - Draw from universal resources to define homology relationships for candidate genes between animal models.
+> - Curate disease classification and meta-analysis through complete CIS listing and relative rank and strength scores.
+> - Offer easy access to external tools from filtered results for deeper biological context.
+
+<div class="notes">
+- relative rank was based either on the number of insertions in a given CIS or the P-value associated with the given CIS
+- The relative rank was calculated by ranking all CISs in a given study and then assigning grades
+- We assigned a letter grade of A for the top 10%, B for the top 11–25%, C for the top 26–50% and D for the bottom 50%.
+</div>
+
+--------------------------------------------------------------------------------
+
+## CCGD Implementation
+
+**ETL Process**
+
+scrape data &rarr; homology association &rarr; integrate into table
+
+![](./fig/search.png)
+
+<div class="container">
+
+<div class="col">
+#### Data Sources/Links
+- NCBI Gene
+- NCBI HomoloGene
+- NCBI PubMed
+- Sanger COSMIC
+- Sanger CGC
+- GeneCards
+</div>
+
+<div class="col">
+> - Filtered results are downloadable.
+> - Wrapped in a simple, instructional website built with classic LAMP stack.
+> - Hosted by UMN-OIT.
+</div>
+
+</div>
+
+<div class="notes">
+- Search Within - mouse human rat fly fish yeast
+</div>
 
@@
 
-# Challenges {.center}
+# Issue {.center}
 
@@
 
-## Forced Upgrades
+## Issue
+
+### Forced OS upgrade from UMN-OIT host
 
@@
 
-## Technical Debt
-> - test
-> - test
+## Challenges
+
+#### Can this application be upgraded as-is?
+
+<div class="notes">
+- several of these items blend limitations that are dependency-based, while others amount to an inconvenience.
+</div>
+
+**Primary Challenges**
+
+> - Machine displacement in provisioning by UMN-OIT
+> - Use of abandonded, proprietary widgets
+> - Outsized SQL database
+
+**Secondary Challenges**
+
+> - Documentation missing visibility for admins
+> - Total app disk use
+> - Lack of version control
+> - Odd hosting paradigm from UMN-OIT
+> - Disorderly codebase, residual WYSIWYG effects
+> - Lack of modularity in back-end
+> - Pending web standards compliance issues
+
+<div class="notes">
+- these are a combination of both limitations and potential goals
+</div>
+
+--------------------------------------------------------------------------------
+
+## Challenges
+
+#### Can this application be upgraded as-is?
+
+**Primary Challenges**
+
+- <span style="color: #a3be8c;">**Machine displacement in provisioning by UMN-OIT**</span>
+- Use of abandonded, proprietary widgets
+- <span style="color: #a3be8c;">**Outsized SQL database**</span>
+
+**Secondary Challenges**
+
+- Documentation missing visibility for admins
+- Total app disk use
+- Lack of version control
+- Odd hosting paradigm from UMN-OIT
+- Disorderly codebase, residual WYSIWYG effects
+- Lack of modularity in back-end
+- Pending web standards compliance issues
+
+--------------------------------------------------------------------------------
+
+## Challenges
+
+### Ultimately, a drop-in transition was not possible.
 
@@
   3   table build       rewrite table build        Rshiny      Rshiny offered a dramatic improvement to replace the existing process by merging the table build back-end with a modern web display of the app interface.
-  4   content update    improved admin controls    BASH/R      Old app version confined some content controls to app author, limiting ability to make contributions by product owner (Tim Starr).
+  4   content update    improved admin controls    BASH/R      Old app version confined some content controls to app author, limiting ability to make contributions by product owner.
   5   version control   implement best practices   git/docs    No version control was used in the original development of the app. This and other documentation practices were expanded in the rewrite.
@@
 
-<div class="column">
+<div class="container">
 
+<div class="col">
 ```{r echo = T}
+---
 title: "Candidate Cancer Gene Database"
@@
 ```
+</div>
 
+<div class="col">
 ```{r echo = T}
+---
 name: "Candidate Cancer Gene Database"
@@
           href: contact.html
-
+---
 ```
+</div>
 
@@
 #### table reactive graph {.center}
+
+<div class="notes">
+- ADD SOME NOTES
+</div>
 
@@
 ## Simplified Codebase
+
+<div class="notes">
+- codebase simplified in a few ways.
+- build table process was dramatically streamlined.
+</div>
+
+```{r echo = T}
+export <- homogTable %>%
+  full_join(df) %>%
+  group_by(MouseId) %>%
+  distinct(Study, .keep_all = T) %>%
+  add_count(name = "Studies") %>%
+  ungroup() %>%
+  mutate(CGC = HumanName %in% cgc) %>%
+  mutate(COSMIC = HumanName %in% cosmic)
+
+write.csv(export, file = "../table_app/ccgd_export.csv", row.names = F)
+
+deployApp(appDir = "../table_app", forceUpdate = T, launch.browser = F)
+```
+
+#### build_table.R table write and app deployment snippet
+
+--------------------------------------------------------------------------------
+
+## Simplified Codebase
+
+<div class="notes">
+- shiny snippet represents full set of code needed to render app ui. app is 200 lines total.
+- lines of code replacement is orders of magnitude. possible 1:100.
+- use of this script allowed for elimination of entire sql database.
+</div>
+
+```{r echo = T}
+export <- homogTable %>%
+  full_join(df) %>%
+  group_by(MouseId) %>%
+  distinct(Study, .keep_all = T) %>%
+  add_count(name = "Studies") %>%
+  ungroup() %>%
+  mutate(CGC = HumanName %in% cgc) %>%
+  mutate(COSMIC = HumanName %in% cosmic)
+
+write.csv(export, file = "../table_app/ccgd_export.csv", row.names = F)
+
+deployApp(appDir = "../table_app", forceUpdate = T, launch.browser = F)
+```
+
+#### build_table.R table write and app deployment snippet
+
 ```{r echo = T}
@@
 
+<div class="notes">
+- Page content is now written in a highly accessible syntax and rendered with a lean script.
+</div>
+
 ```{r echo = T}
@@
 #### index.html page content snippet {.center}
+
+--------------------------------------------------------------------------------
+
+## Simplified Codebase
+
+```{r echo = T}
+---
+title: "Candidate Cancer Gene Database"
+bibliography: refs/ccgd_paper.bib
+nocite: '@*'
+output:
+    html_document:
+        includes:
+            in_header: "styles/favicon.html"
+            after_body: "styles/footer.html"
+        css: styles/styles.css
+        theme: readable
+---
+
+The Candidate Cancer Gene Database (CCGD) was developed to make accessible a collated set of results from transposon-based forward cancer genetic screens in mice.
+The value in doing this is to give investigators the ability to quickly filter through the results of many such screens in an effort to determine the candidacy of a gene for its role in cancer.
+
+***
+
+### Background
+
+This work is a product of the [Starr Lab](http://www.med.umn.edu/starrlab/) at the [University of Minnesota](http://www.umn.edu/).
+The Starr Lab is affiliated with the [Center for Genome Engineering](http://www.cge.umn.edu/), the [Masonic Cancer Center](http://www.cancer.umn.edu/), the Department of [OB/GYN](http://www.obgyn.umn.edu/), and the Department of [Genetics, Cell Biology, and Development](http://www.cbs.umn.edu/gcd/).
+
+##### Contributions
+
+The idea for creating this database originated from Tim Starr, Vincent Keng, and David Largaespada at UMN.
+Erik Nyre and Ken Abbott developed the original database, with help from Juan Abrahante.
+A complete website and backend overhaul was conducted by Christopher Tastad in 2019.
+
+* Ken Abbott
+* [Erik Nyre](https://www.dorsey.com/people/n/nyre-erik)
+* [Juan Abrahante](https://research.umn.edu/units/umii/about-us/staff-directory/juan-e-abrahante-phd)
+* [Christopher Tastad](https://github.com/ctastad)
+
+***
+
+##### Updates
+
+The most recent table build was completed on:
+
+<object type="text/html"
+        height=40px
+        data="build_date.txt"></object>
+
+```
+
+#### index.html page content snippet {.center}
+
+```{r echo = T}
+setwd("..")
+
+# render core webpages
+rmarkdown::render("index.Rmd")
+rmarkdown::render("search.Rmd")
+rmarkdown::render("help.Rmd")
+rmarkdown::render("references.Rmd")
+rmarkdown::render("contact.Rmd")
+rmarkdown::render_site("index.Rmd")
+```
+
+#### knit_site.R page and site rendering script
 
@@
 
+--------------------------------------------------------------------------------
+
 ## Conclusions
+- Stated requirements set out by this project have been met.
+- A new, more modern and simplified framework has been implemented across the full application stack.
+- All original functionality has been sustained.
+- Design and engineering considerations were given to ease future maintenance.
+
+## Future Directions
+- Continuous integration/continuous deployment
+- Anticipating future upgrades and framework evolution
+
+--------------------------------------------------------------------------------
 
 ## Acknowledgements
+
+--------------------------------------------------------------------------------
 
@@
 
-.column {
-  column-count: 2;
+.container{
+    display: flex;
+}
+
+.col{
+    flex: 1;
 }

### 2021-05-21 presentation.Rmd
final edits

@@
 
-> - Cancer bio is complicated.
-> - Driver genes present primary targets.
+> - Cancer bio is complicated
+> - Driver genes present primary targets
 > - Screening methods...
@@
 - what purpose does it offer the field
+
+---
 
@@
 
-> - Create an interactive web tool to quickly refine the stated data set by a variety of parameters.
-> - Draw from universal resources to define homology relationships for candidate genes between animal models.
-> - Curate disease classification and meta-analysis through complete CIS listing and relative rank and strength scores.
-> - Offer easy access to external tools from filtered results for deeper biological context.
+> - Create a web tool
+> - Form homology relationships
+> - Curate disease classification
+> - Offer connections to external tools
 
 <div class="notes">
+- Create an interactive web tool to quickly refine the stated data set by a variety of parameters.
+- Draw from universal resources to define homology relationships for candidate genes between animal models.
+- Curate disease classification and meta-analysis through complete CIS listing and relative rank and strength scores.
+- Offer easy access to external tools from filtered results for deeper biological context.
+
+---
+
 - relative rank was based either on the number of insertions in a given CIS or the P-value associated with the given CIS
@@
 <div class="col">
-> - Filtered results are downloadable.
-> - Wrapped in a simple, instructional website built with classic LAMP stack.
-> - Hosted by UMN-OIT.
+> - Filtered results are downloadable
+> - Wrapped in a simple, instructional website built with classic LAMP stack
+> - Hosted by UMN-OIT
 </div>
@@
 > - Machine displacement in provisioning by UMN-OIT
-> - Use of abandonded, proprietary widgets
+> - Use of abandoned, proprietary widgets
 > - Outsized SQL database
@@
 - <span style="color: #a3be8c;">**Machine displacement in provisioning by UMN-OIT**</span>
-- Use of abandonded, proprietary widgets
+- Use of abandoned, proprietary widgets
 - <span style="color: #a3be8c;">**Outsized SQL database**</span>
@@
 <div class="notes">
-- ADD SOME NOTES
+- rshiny format is ui + server -> deployment
+- server is made up of reactive constructs of inputs, expressions, and outputs
+- employs declarative programming style allowing for lazy code execution.
+- define loose, high level controls to passively fill functional requirements
+- improves flexibility for dev and machine - resource use reduce through modularity in code execution
+- our app uses 2 state path
 </div>
@@
 #### server-side processing flow {.center}
+
+<div class="notes">
+- typically, server side processing with a table of this side would be prohibitive
+</div>
 
@@
 - build table process was dramatically streamlined.
-</div>
-
-```{r echo = T}
-export <- homogTable %>%
-  full_join(df) %>%
-  group_by(MouseId) %>%
-  distinct(Study, .keep_all = T) %>%
-  add_count(name = "Studies") %>%
-  ungroup() %>%
-  mutate(CGC = HumanName %in% cgc) %>%
-  mutate(COSMIC = HumanName %in% cosmic)
-
-write.csv(export, file = "../table_app/ccgd_export.csv", row.names = F)
-
-deployApp(appDir = "../table_app", forceUpdate = T, launch.browser = F)
-```
-
-#### build_table.R table write and app deployment snippet
-
---------------------------------------------------------------------------------
-
-## Simplified Codebase
-
-<div class="notes">
 - shiny snippet represents full set of code needed to render app ui. app is 200 lines total.
-- lines of code replacement is orders of magnitude. possible 1:100.
+- lines of code replacement is orders of magnitude. possibly 1:100.
 - use of this script allowed for elimination of entire sql database.
 </div>
-
-```{r echo = T}
-export <- homogTable %>%
-  full_join(df) %>%
-  group_by(MouseId) %>%
-  distinct(Study, .keep_all = T) %>%
-  add_count(name = "Studies") %>%
-  ungroup() %>%
-  mutate(CGC = HumanName %in% cgc) %>%
-  mutate(COSMIC = HumanName %in% cosmic)
-
-write.csv(export, file = "../table_app/ccgd_export.csv", row.names = F)
-
-deployApp(appDir = "../table_app", forceUpdate = T, launch.browser = F)
-```
-
-#### build_table.R table write and app deployment snippet
 
@@
 ## Conclusions
-- Stated requirements set out by this project have been met.
-- A new, more modern and simplified framework has been implemented across the full application stack.
-- All original functionality has been sustained.
-- Design and engineering considerations were given to ease future maintenance.
+> - Stated requirements set out by this project have been met.
+> - A new, more modern and simplified framework has been implemented across the full application stack.
+> - All original functionality has been sustained.
+> - Design and engineering considerations were given to ease future maintenance.
+
+--------------------------------------------------------------------------------
 
 ## Future Directions
-- Continuous integration/continuous deployment
-- Anticipating future upgrades and framework evolution
+> - Continuous integration/continuous deployment
+> - Anticipating future upgrades and framework evolution
 
@@
 
+- Tim Starr*
+- Yuk Sham*
+- Jinhua Wang
+- National Marrow Donor Program
+- UMN-OIT
+- Cassandra Price
+- My wife and parents
+
 --------------------------------------------------------------------------------
@@
 ## The End {.center}
+
+--------------------------------------------------------------------------------
+
+# Supplemental
+
+http://hst-ccgd-prd-web.oit.umn.edu/search.html
 

### 2021-05-21 presentation.Rmd
final

@@
 - Jinhua Wang
+- Ken Abbott
+- BICB program
 - National Marrow Donor Program
 - UMN-OIT
+- Judy Cho
+- Phil Schumm
 - Cassandra Price
+- Nouf Gazdar
 - My wife and parents
